\documentclass[10pt]{beamer}
%\usepackage[whole]{bxcjkjatype}

\usepackage{newtxtext,newtxmath}
\usepackage{stmaryrd}
\usepackage{dsfont}
\usepackage{xfrac}
\usepackage{mathrsfs}

\usepackage{booktabs}
\usepackage{multirow}

\usepackage[utf8]{inputenc}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage[noend]{algpseudocode}
\usepackage{algpseudocode}
\usepackage{multicol}

\usepackage{arydshln}
\usepackage{here}

\usepackage{graphics}
\usepackage{amsmath}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\newcommand{\argmin}{\mathop{\rm arg~min}}

\newtheorem{proposition}{Proposition}
\newtheorem*{threatmodel}{Threat Model}
\newtheorem{scenario}{Scenario}

\newcommand{\hquad}{\hspace{0.5em}}

\input{settings}
\usetikzlibrary{positioning,arrows.meta,calc,fit,backgrounds,shapes.geometric}

% ----- math shortcuts (kept consistent with the paper) -----
\newcommand{\FF}{\mathbb{F}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bO}{\mathbf{O}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bK}{\mathbf{K}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\softmax}{\mathrm{softmax}}
\DeclareMathOperator{\LN}{LN}
\DeclareMathOperator{\LLA}{LLA}
\DeclareMathOperator{\Att}{Att}
\DeclareMathOperator{\Com}{\mathsf{Com}}
\newcommand{\sys}{\ensuremath{\pi\text{-Former}}}

\title[$\pi$-Former]{$\pi$-Former: Succinct Proofs of Transformer Inference\\
via Structured Lookup Attention}
\author[H. Takahashi]{Hideaki Takahashi}
\date[]{April 2026}

\begin{document}
\Front

\pagestyle{empty}

%======================================================================
\begin{frame}{What Is a Transformer? (Intuition)}
A transformer is a stack of $L$ identical \textbf{blocks}.
Each block takes a sequence of $T$ token vectors
$\bX\in\RR^{T\times d}$ and returns the same shape.

\medskip
Inside one block, two operations:
\begin{enumerate}
  \item \textbf{Attention}: each token looks at every other token
        and pulls in information from the relevant ones.
  \item \textbf{Feed-forward (FFN)}: each token is independently
        passed through a small two-layer MLP.
\end{enumerate}

\medskip
Each is wrapped with: \textbf{LayerNorm} (rescale the row),
a \textbf{residual} ``$+$'' (add the input back).

\medskip
\begin{center}
\small
$\bX_{\mathrm{mid}} = \bX_{\mathrm{in}} + \Att(\LN(\bX_{\mathrm{in}}))$
\hspace{1em}
$\bX_{\mathrm{out}} = \bX_{\mathrm{mid}} + \mathrm{FFN}(\LN(\bX_{\mathrm{mid}}))$
\end{center}

\medskip
That's it. Repeat $L$ times, project to vocabulary, predict next token.
\end{frame}

%======================================================================
\begin{frame}{What Is Attention? (Intuition)}
Each token is given three vectors:
\begin{itemize}
  \item \textbf{Query} $\bq$ -- ``what am I looking for?''
  \item \textbf{Key} $\bk$ -- ``what do I offer?''
  \item \textbf{Value} $\bv$ -- ``what info do I carry?''
\end{itemize}
all linear projections of the input row: $\bq=\bx\bW_Q$, etc.

\medskip
For token $t$, attention computes a \emph{weighted average of values},
where the weight is how well its query matches each key:
\[
  \mathrm{out}_t \;=\; \sum_{s} \underbrace{\softmax_s\!\bigl(\langle\bq_t,\bk_s\rangle/\sqrt d\bigr)}_{\text{attention weight}}\;\bv_s.
\]

\medskip
\textbf{What this gives the model.} Tokens can ``ask'' for
information from other positions and aggregate it -- no fixed
window, no recurrence.

\medskip
\textbf{What this costs the SNARK.} Softmax means
$\exp(\cdot)$ and row-wise division -- both \emph{very} expensive
in a finite field. This is what we will replace.
\end{frame}

%======================================================================
\begin{frame}{Motivation \& Challenge}
\textbf{Setting.} A model owner runs $\mathcal{M}_\theta$ on user input
$\bx$ and returns $\by$. The user wants a cryptographic proof that
$\by=\mathcal{M}_\theta(\bx)$, with short proof and cheap verification.

\medskip
\textbf{The catch.} A SNARK only natively expresses
\textbf{additions and multiplications in $\FF$}. But a transformer block has:
\begin{itemize}
  \item \textbf{Softmax}: $\exp(\cdot)$ + row-wise division.
  \item \textbf{GeLU}: transcendental.
  \item \textbf{LayerNorm}: $\sqrt{\sigma^2+\epsilon}$ + division.
  \item \textbf{Dense linear}: one field multiplication per weight.
\end{itemize}

\medskip
State of the art (zkLLM, zkGPT) approximates these with large lookup
tables. Prover time still \textbf{minutes per token} for GPT-scale models.

\medskip
\begin{block}{Our angle}
Don't fight the model. \textbf{Co-design} the model and the proof
system so that every op is natively cheap.
\end{block}
\end{frame}

%======================================================================
\begin{frame}{Co-Design in One Slide}
We replace each painful operation with a SNARK-friendly twin:

\medskip
\begin{center}
\begin{tabular}{l@{\hskip 0.6em}c@{\hskip 0.6em}l}
\toprule
\textbf{Standard op} & $\Rightarrow$ & \textbf{$\pi$-Former op} \\
\midrule
softmax attention   & $\Rightarrow$ & \textbf{LLA}: linear attention with lookup $\bphi$ \\
GeLU                & $\Rightarrow$ & learnable additive sub-table lookup \\
LayerNorm (div, $\sqrt{\,}$) & $\Rightarrow$ & constraint-fused LN, no div / $\sqrt{\,}$ \\
dense $\bW$ (FP)    & $\Rightarrow$ & ternary $\bW\in\{-1,0,+1\}\cdot\alpha$ \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{End result.} Every forward-pass operation becomes
\begin{itemize}
  \item a low-degree \textbf{sumcheck} claim, or
  \item a \textbf{structured table lookup} (Lasso).
\end{itemize}

\medskip
No ``ZK-unfriendly operation tax'' to pay.
\end{frame}

%======================================================================
\begin{frame}{Toolbox (One-Line Reminders)}
\textbf{Sumcheck.} Reduces $H=\sum_{x\in\{0,1\}^n} f(x)$ to a single
evaluation $f(r)$ at random $r$. Verifier cost $O(n\delta)$,
soundness error $\le n\delta/|\FF|$.

\medskip
\textbf{Hyrax PCS.} Transparent (no trusted setup) polynomial
commitment based on Pedersen / DL on BN254 G1. $O(\sqrt N)$
commitment size, $O(\sqrt N)$ open. Importantly: \textbf{additively
homomorphic}.

\medskip
\textbf{Lasso.} A lookup argument that proves
$v_j = T[\mathrm{idx}_j]$ for many queries against a table $T$,
\emph{cheap when $T$ decomposes into small additive sub-tables}.

\medskip
\begin{block}{Strategy}
Build the model so that:
attention $=$ sumcheck $+$ Lasso,
activation $=$ Lasso,
LayerNorm $=$ sumchecks $+$ range checks,
projection $=$ sumcheck with no field mults.
\end{block}
\end{frame}

%======================================================================
\begin{frame}{The $\pi$-Former Block}
\begin{center}
\begin{tikzpicture}[
  font=\scriptsize,
  box/.style={rectangle, draw, rounded corners=2pt, minimum width=1.2cm,
              minimum height=0.5cm, align=center, fill=blue!6},
  op/.style={rectangle, draw, rounded corners=2pt, minimum width=1.2cm,
             minimum height=0.5cm, align=center, fill=green!10},
  add/.style={circle, draw, inner sep=0.5pt, font=\tiny, fill=white},
  arrow/.style={-latex, thick}
]
  \node[box] (xin)        at (0,0)   {$\bX_{\mathrm{in}}$};
  \node[op,  right=0.25cm of xin]  (ln1)  {$\LN_1$};
  \node[op,  right=0.25cm of ln1]  (qkv)  {QKV\\(ternary)};
  \node[op,  right=0.25cm of qkv]  (lla)  {LLA\\(Lasso $\bphi$)};
  \node[op,  right=0.25cm of lla]  (woo)  {$\bW_O$\\(ternary)};
  \node[add, right=0.25cm of woo]  (res1) {$+$};
  \node[box, right=0.25cm of res1] (xmid) {$\bX_{\mathrm{mid}}$};

  \draw[arrow] (xin) -- (ln1);
  \draw[arrow] (ln1) -- (qkv);
  \draw[arrow] (qkv) -- (lla);
  \draw[arrow] (lla) -- (woo);
  \draw[arrow] (woo) -- (res1);
  \draw[arrow] (res1) -- (xmid);
  \draw[arrow] (xin.north) |- ($(xin.north)+(0,0.45)$) -| (res1.north);

  \node[op,  below=0.6cm of ln1]  (ln2)  {$\LN_2$};
  \node[op,  right=0.25cm of ln2] (w1)   {$\bW_1$\\(ternary)};
  \node[op,  right=0.25cm of w1]  (phi)  {$\bphi$\\(Lasso)};
  \node[op,  right=0.25cm of phi] (w2)   {$\bW_2$\\(ternary)};
  \node[add, right=0.25cm of w2]  (res2) {$+$};
  \node[box, right=0.25cm of res2] (xout) {$\bX_{\mathrm{out}}$};

  \draw[arrow] (xmid.south) |- (ln2.west);
  \draw[arrow] (ln2) -- (w1);
  \draw[arrow] (w1) -- (phi);
  \draw[arrow] (phi) -- (w2);
  \draw[arrow] (w2) -- (res2);
  \draw[arrow] (res2) -- (xout);
  \draw[arrow] (xmid.south west) |- ($(ln2.north)+(0,0.35)$) -| (res2.north);
\end{tikzpicture}
\end{center}

\textbf{Color code.}
\begin{itemize}
  \item Green = an op the SNARK proves.
  \item Ternary boxes have \emph{no} field multiplications inside.
  \item LLA / $\bphi$ are proved via Lasso lookups.
  \item ``$+$'' (residual) is checked by \textbf{homomorphic addition
        of Hyrax commitments} -- no proof at all.
\end{itemize}
\end{frame}

%======================================================================
\begin{frame}{Idea 1: Learnable Lookup Attention (LLA)}
\textbf{Linear attention} replaces $\softmax(\bQ\bK^\top/\sqrt{d})$ by
a positive-definite kernel
$\kappa(\bq,\bk)=\langle\bphi(\bq),\bphi(\bk)\rangle$:
\[
  \LLA(\bQ,\bK,\bV)_t \;=\;
  \bphi(\bq_t)^{\!\top}\,\underbrace{\sum_{s}\bphi(\bk_s)\bv_s^{\!\top}}_{=\,\bM\;(\text{context, computed once})}.
\]
Cost: $O(n\,d^2)$ instead of $O(n^2 d)$, no exp, no row division.

\medskip
\textbf{Where's $\bphi$?} We make it learnable and SNARK-friendly:
\[
  \bphi(x) \;=\; \sum_{i=0}^{c-1} T_i\bigl[\chi_i(x_{\mathrm{int}})\bigr],
\]
i.e.\ split a $b$-bit input into $c$ chunks, look each up in a
small table $T_i$ of size $2^{b/c}$, sum.

\medskip
\textbf{Why this matters.}
\begin{itemize}
  \item For $b{=}16,\,c{=}2$: two 256-entry tables instead of one
        65{,}536-entry table.
  \item This is \emph{exactly} the structured-decomposability
        condition Lasso wants.
  \item Init so $\sum_i T_i \approx \mathrm{GeLU}$, then train end-to-end
        with the straight-through estimator.
\end{itemize}
\end{frame}

%======================================================================
\begin{frame}{Idea 2: Ternary Projections}
All projection matrices ($\bW_Q,\bW_K,\bW_V,\bW_O,\bW_1,\bW_2$, LM head)
are constrained:
\[
  \bW \in \{-1, 0, +1\}^{m\times n},\qquad
  \by = \alpha\cdot(\bx\bW) + \mathbf{b}
\]
with one learnable scalar $\alpha$ per layer (BitNet b1.58 style).

\medskip
\textbf{What changes inside the SNARK?} Consider the standard
sumcheck for a contraction $\bY = \bX\bW$:
\[
  \widetilde\bY(r_t,r_d) - \mathbf{b}(r_d)
  \;=\; \alpha \sum_k \widetilde\bX(r_t,k)\,\widetilde\bW(k,r_d).
\]
Because $\widetilde\bW(k,r_d)\in\{-1,0,+1\}$ after evaluation,
each prover round is:
\begin{center}
``$+\!=$ / $-\!=$ / skip'' \quad -- no field multiplications.
\end{center}

\medskip
\begin{block}{Bonus}
The verifying key never holds raw weights -- only
Hyrax \emph{commitments} to the ternary arrays.
\end{block}
\end{frame}

%======================================================================
\begin{frame}{Idea 3: Constraint-Fused LayerNorm}
Standard LN has $\sqrt{\sigma^2+\epsilon}$ and a division --
both bad in $\FF$. We rewrite the identity as integer relations:

\medskip
For each row, audit at \emph{one} random index $r_t$ via four checks:
\begin{enumerate}
  \item \textbf{Mean}: degree-2 sumcheck for $\widetilde{\mathrm{sum\_x}}(r_t)$.
  \item \textbf{Variance}: cubic sumcheck on the centered residual.
  \item \textbf{Sigma range}: prove $\sigma=\lfloor\sqrt{\mathrm{var}}/d\rfloor$
        by range-checking
        $\mathrm{var}-(d\sigma)^2 \ge 0$ and
        $(d\sigma+d)^2-1-\mathrm{var}\ge 0$.
  \item \textbf{$\bY$ fusion}: a single multi-batched cubic sumcheck
        for the $\gamma\bX$ and $\sigma\bY$ legs together.
\end{enumerate}

\medskip
\textbf{Result.} No field divisions, no square roots. Verifier reads
$\gamma,\beta$ from the VK; everything else is one Hyrax opening per
witness.
\end{frame}

%======================================================================
\begin{frame}{Cross-Block Batching: The Key Systems Trick}
A transformer is the \emph{same protocol, run $L$ times}.
So instead of $L$ independent proofs per protocol type, we batch:

\medskip
\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Resource} & \textbf{Folded across blocks into\dots} \\
\midrule
6 sumcheck types  & 6 \emph{model-level} sumchecks \\
all Lasso instances     & one global multi-Lasso \\
all LN range proofs     & one shared multiplicity commitment \\
residual $+$            & no proof; just add Hyrax commitments \\
final openings          & 10 deferred Hyrax accumulators \\
                        & finalised in 1 MSM pair each \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{The fold.} For each protocol type, after the verifier sees
all per-block claims $H_\ell$, draw $\eta\leftarrow\mathsf{FS}$ and
prove
\[
  H^\star \;=\; \sum_{\ell=1}^L \eta^{\ell-1} H_\ell
\]
with one inner sumcheck. Standard random-linear-combination argument
$\Rightarrow$ extra error only $(L-1)/|\FF|$.
\end{frame}

%======================================================================
\begin{frame}{End-to-End Performance}
Single-thread x86\_64 CPU. Times are pure compute; IO excluded.

\medskip
\begin{center}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lrrrrrrrr}
\toprule
\textbf{Model} & \textbf{$d$} & \textbf{$L$} & \textbf{$d_{ff}$} & \textbf{$T$}
   & \textbf{Params} & \textbf{Prove} & \textbf{Verify} & \textbf{Proof} \\
\midrule
tiny   & 8   & 1 & 16  & 8  & 1.1k  & 0.22\,s & 0.03\,s & 108\,KB \\
small  & 64  & 2 & 256 & 16 & 133k  & 1.10\,s & 0.15\,s & 1.0\,MB \\
medium & 128 & 4 & 512 & 32 & 924k  & \textbf{4.24\,s} & \textbf{0.28\,s} & 5.5\,MB \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{Headline.} 4-layer, $\sim$924k-parameter transformer:
\textbf{4.2\,s prove, 280\,ms verify, 5.5\,MB proof} on a single CPU core.

\medskip
\textbf{Scaling.}
\begin{itemize}
  \item Prove time $\sim$ linear in $L\cdot T\cdot d$ (per-block
        arithmetic survives batching).
  \item Verify time \emph{sub-linear}: each cross-block sumcheck is
        a $\log Td$-round transcript, accumulators finalise in 1 MSM
        pair each.
\end{itemize}
\end{frame}

%======================================================================
\begin{frame}{Where Does the Verifier Spend Time?}
\begin{center}
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lrrr}
\toprule
\textbf{Phase} & \textbf{tiny} & \textbf{small} & \textbf{medium} \\
\midrule
6 cross-block sumchecks (combined) & $<$0.2 & $<$0.5 & $<$0.8 \\
Final range / final LN / LM-head   & 1.7    & 1.9    & 3.9   \\
Cross-block batched opens          & 3.6    & 44.5   & 46.6  \\
Global Lasso                       & 1.6    & 25.6   & 40.9  \\
Accumulator finalisation (10 MSMs) & 13.2   & 23.7   & 54.2  \\
\midrule
\textbf{Subtotal (ms)}             & 22     & 99     & 148   \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{Take-aways.}
\begin{itemize}
  \item Cross-block sumcheck transcripts are \emph{negligible}
        ($<\!1$\,ms) -- batching collapses transcript work.
  \item Hyrax openings + accumulator MSMs dominate.
  \item These are exactly the inner loops that GPU MSM /
        parallel sumcheck accelerate.
\end{itemize}
\end{frame}

%======================================================================
\begin{frame}{Comparison, Limitations, Roadmap}
\begin{center}
\scriptsize
\begin{tabular}{lll}
\toprule
\textbf{System} & \textbf{Attention} & \textbf{Activation} \\
\midrule
zkLLM   & softmax (\texttt{tlookup}) & structured lookup \\
zkGPT   & softmax (approx., Plonky2) & lookup table \\
\sys{}  & \textbf{linear (LLA, exact)} & \textbf{learnable} additive lookup \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Position.} \sys{} attention is not a softmax approximation:
it is a different (kernelised) attention trained from scratch, so
the SNARK proves the model \emph{exactly}. $\bphi$ is learnable.

\medskip
\textbf{Currently in the SNARK.} Single-head; un-normalised LLA
numerator; no causal mask; \emph{not} zero-knowledge.

\medskip
\textbf{Engineering, not architecture.}
\begin{itemize}
  \item Multi-head: parallel batches of the same sumcheck.
  \item Causal mask: an indicator polynomial inside the sumcheck.
  \item ZK: standard blinding layer over committed witnesses.
  \item GPU MSM + parallel sumcheck: every dominant inner loop is
        data-parallel.
  \item Streaming generation: Nova-style folding with
        $\bM=\bphi(\bK)^\top\bV$ as the IVC accumulator
        $\Rightarrow$ constant amortised cost per token.
\end{itemize}
\end{frame}

%======================================================================
\begin{frame}{Take-aways}
\begin{enumerate}
  \item \textbf{Co-design beats approximation.}
        Replace softmax / GeLU / LN / dense $\bW$ with twins that
        are \emph{natively} sumcheck $+$ Lasso.
  \item \textbf{LLA + ternary + fused LN}: every transformer op has
        sub-linear verifier cost.
  \item \textbf{Cross-block batching} folds $L$ per-block proofs into
        6 model-level sumchecks, 1 global Lasso, 1 global range,
        10 deferred openings.
  \item \textbf{Result.} 4-layer, 924k-parameter transformer:
        \textbf{4.2\,s prove / 280\,ms verify / 5.5\,MB proof},
        single CPU core, transparent setup.
\end{enumerate}

\bigskip
\begin{center}
\Large Thank you.\\
\medskip
\small \texttt{ht2673@columbia.edu}
\end{center}
\end{frame}

\end{document}
