%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \pi-Former: Succinct Proofs of Transformer Inference
% USENIX submission template (uses usenix-2020-09.sty).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{usenix-2020-09}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{bbm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,calc,fit,backgrounds,shapes.geometric}
\usepackage{enumitem}
\setlist[itemize]{leftmargin=1.4em,topsep=2pt,itemsep=2pt}
\setlist[enumerate]{leftmargin=1.6em,topsep=2pt,itemsep=2pt}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\algrenewcommand\algorithmicrequire{\textbf{Input:}}
\algrenewcommand\algorithmicensure{\textbf{Output:}}
\algnewcommand{\LineComment}[1]{\Statex \(\triangleright\) \emph{#1}}
\algnewcommand{\Phase}[1]{\Statex \textbf{Phase #1}}

\newcommand{\sys}{\ensuremath{\pi\text{-Former}}}
\newcommand{\FF}{\mathbb{F}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\bq}{\mathbf{q}}
\newcommand{\bk}{\mathbf{k}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bo}{\mathbf{o}}
\newcommand{\bO}{\mathbf{O}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bb}{\mathbf{b}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bN}{\mathbf{N}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bK}{\mathbf{K}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\calM}{\mathcal{M}}
\newcommand{\calT}{\mathcal{T}}
\newcommand{\calP}{\mathcal{P}}
\newcommand{\calV}{\mathcal{V}}
\newcommand{\calA}{\mathcal{A}}
\newcommand{\piproof}{\pi}
\newcommand{\bits}{B}
\newcommand{\chunks}{c}
\newcommand{\chunkbits}{m}
\newcommand{\polylog}{\mathrm{polylog}}
\newcommand{\poly}{\mathrm{poly}}
\newcommand{\softmax}{\mathrm{softmax}}
\newcommand{\eq}{\widetilde{\mathrm{eq}}}
\DeclareMathOperator{\LN}{LN}
\DeclareMathOperator{\GELU}{GELU}
\DeclareMathOperator{\Att}{Att}
\DeclareMathOperator{\Com}{\mathsf{Com}}
\DeclareMathOperator{\Open}{\mathsf{Open}}
\DeclareMathOperator{\Verify}{\mathsf{Verify}}
\DeclareMathOperator{\Prove}{\mathsf{Prove}}
\DeclareMathOperator{\Setup}{\mathsf{Setup}}
\DeclareMathOperator{\poppk}{\mathsf{pk}}
\DeclareMathOperator{\popvk}{\mathsf{vk}}

%don't want date printed
\date{}

\title{\Large \bf \sys{}: Succinct Proofs of Transformer Inference\\
       via a SNARK-Friendly Architecture}

\author{
{\rm Hideaki Takahashi}\\
Columbia University\\
\texttt{ht2673@columbia.edu}
}

\begin{document}

\maketitle

%-------------------------------------------------------------------------------
\begin{abstract}
%-------------------------------------------------------------------------------
Verifiable transformer inference lets an untrusted operator prove
that a posted output was produced by a committed model, without
revealing the weights. The obstacle is that a transformer block is
built from exactly the operations that finite-field SNARKs find
expensive: softmax, GeLU, LayerNorm, and dense projections. Prior
systems leave the model untouched and pay this cost on every block.

We present \sys{}, a succinct argument whose central design choice
is to co-design the \emph{model} so the proof system has nothing
expensive to absorb. Three architectural changes drive the design:
(i) softmax attention is replaced by a kernel attention whose feature
map is realised as an additively decomposed learnable lookup, so the
only non-arithmetic operation is a structured Lasso table look-up;
(ii) every projection is constrained to ternary entries scaled by a
per-layer scalar, eliminating all field multiplications from weight
contractions; and (iii) LayerNorm is reformulated as an integer
protocol provable through sumcheck and chunked range proofs. We build
a transparent BN254 SNARK on top with Hyrax commitments, batching
every per-block sub-protocol into a constant number of model-level
sumchecks, a globally shared range-multiplicity commitment, and a
zerocheck-folded inverse polynomial that removes $B\!\cdot\!C$ Hyrax
commitments per range bucket. Residual connections require no proof:
they are verified by linearity of the commitment scheme. We prove
end-to-end soundness under discrete log on BN254 G1 and implement a
complete prototype in $\sim$19k lines of Rust with a bit-identical
PyTorch training pipeline.
\end{abstract}

%-------------------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}
%-------------------------------------------------------------------------------

Large language models are increasingly deployed as black-box
services. The user submits a prompt, an opaque operator runs a
proprietary model, and the user has no cryptographic guarantee that
the returned output came from the advertised model. Verifiable
inference~\cite{liu2021zkcnn,sun2024zkllm,qu2025zkgpt,abbaszadeh2024zero}
closes this gap with a succinct non-interactive argument: the
operator commits to the weights once, and for every input produces a
small proof that the verifier checks without re-executing the model
and without learning the weights.

The core difficulty is that a transformer block is built from
operations that finite-field SNARKs find expensive. Softmax requires
exponentials and a row-wise division; GeLU is transcendental;
LayerNorm contains a square root and a division by a running
variance; a dense projection costs one field multiplication per
weight entry. Prior work attacks this on the proof-system side:
zkLLM~\cite{sun2024zkllm} introduces \texttt{tlookup}, a parallelised
lookup argument tailored to softmax and SwiGLU; zkGPT~\cite{qu2025zkgpt}
pairs Plonky2 with constraint fusion to drive down the cost of
approximating softmax. Both keep the model fixed and pay an
``unfriendly-operation tax'' on every block.

\paragraph{Our approach.}
\sys{} starts from a different premise: change the \emph{model} so
that the proof system has nothing expensive to absorb. Three
architectural principles drive the design:
\begin{itemize}
  \item \textbf{SNARK-friendly operators.} Replace every operation
        without a compact arithmetic representation. Softmax becomes
        a kernel attention whose only non-arithmetic component is a
        structured table lookup; LayerNorm becomes an integer
        protocol provable by sumcheck plus chunked range proofs.
  \item \textbf{Ternary projections for free contractions.}
        Constrain every projection matrix to entries
        $\{-1,0,+1\}$ scaled by a per-layer scalar
        $\alpha$~\cite{ma2025bitnet}. A degree-$2$ sumcheck against
        a ternary weight folds the equality polynomial against
        ternary entries with $+\!=$/$-\!=$/skip; no field
        multiplications appear in the contraction.
  \item \textbf{Learnable lookup tables.} The element-wise
        non-linearity $\bphi$ is itself learnable: it is parametrised
        as a sum of small sub-tables and trained end-to-end. Under
        Lasso~\cite{setty2024unlocking}, prover cost depends only on
        table size and query count, not on table contents, so
        training adapts table values at zero proving cost.
\end{itemize}
Together these reduce a transformer block to operations that map
directly to a low-degree sumcheck or to a Lasso lookup over a small
structured table.

\paragraph{Model-level batching.}
On top of this architecture we build a proof system around a single
observation: the same protocol runs once per block with independent
inputs but identical structure. \sys{} runs six model-level
sumchecks (one per protocol type) that fold all $L$ per-block claims
via a Fiat--Shamir random linear combination; the inner sumcheck
transcript is shared across blocks. The same idea applies to Lasso
(one global multi-instance proof per lookup family), to range checks
(one shared multiplicity commitment per width bucket spanning the
entire model), to residuals (homomorphic Hyrax addition; no proof at
all), and to the final opening stage (deferred batch accumulators
finalised in one MSM pair per parameter group). For the range proof
specifically, we replace the explicit Hyrax commitment to each LogUp
inverse polynomial by a zerocheck folded into the bucket sumcheck;
this removes $B\cdot C$ commitments per bucket without changing the
per-round prover cost.

\paragraph{Contributions.}
\begin{enumerate}
  \item A SNARK-friendly transformer block
        (\S\ref{sec:arch}) in which attention, normalisation, and
        projection each map to a sumcheck or a Lasso lookup, combining
        learnable lookup attention, ternary projections, and a
        constraint-fused integer LayerNorm.
  \item A complete protocol (\S\ref{sec:protocol}) batching all
        $L$ per-block sub-proofs into six model-level cross-block
        sumchecks, one global Lasso for the $\bphi$ family, and a
        bucket-batched range proof shared across the whole model.
  \item A zerocheck-folded LogUp range proof (\S\ref{ssec:range},
        Theorem~\ref{thm:range}) that eliminates $B\cdot C$
        Hyrax commitments per width bucket while keeping the
        bucket sumcheck cubic and soundness within $O(n/|\FF|)$.
  \item End-to-end soundness (\S\ref{sec:soundness},
        Theorem~\ref{thm:end-to-end}) reducing to discrete log on
        BN254 G1, and a working prototype in $\sim$19k Rust LoC with
        a bit-identical PyTorch training pipeline
        (\S\ref{sec:impl}).
\end{enumerate}

\paragraph{Threat model.}
We target the standard non-interactive argument-of-knowledge setting
in the random-oracle model. The verifier holds a verifying key
$\popvk$ binding the weights and lookup tables, a public input
$\bx_{\mathrm{in}}$, and a claimed output $\by$, and accepts iff
$\calM_\theta(\bx_{\mathrm{in}})\!=\!\by$ for the committed model.
Soundness is computational under discrete log on BN254 G1, the
standard assumption for the Hyrax PCS~\cite{wahby2018doubly}. The
current implementation is not zero-knowledge: weight commitments are
public and intermediate sumcheck evaluations are revealed. The
standard masking extension is discussed in \S\ref{sec:disc}.

%-------------------------------------------------------------------------------
\section{Preliminaries}
\label{sec:bg}
%-------------------------------------------------------------------------------

\paragraph{Notation.}
We work over the BN254 scalar field $\FF$ ($r\!\approx\!2^{254}$) and
the BN254 G1 group $\GG$. We use $L$ for the number of blocks, $T$
for the sequence length, $d$ for the embedding dimension, and $d_{ff}$
for the FFN width; the prototype is single-head ($d_h\!=\!d$). A
multilinear polynomial $\widetilde f$ on $n$ variables is identified
with its $2^n$ evaluations on $\{0,1\}^n$, and the equality polynomial
$\eq(r,x)\!=\!\prod_i(r_i x_i+(1\!-\!r_i)(1\!-\!x_i))$ satisfies
$\sum_x\eq(r,x)\widetilde f(x)=\widetilde f(r)$. Real-valued tensors are
quantised to integers and embedded in $\FF$ before commitment; our
training pipeline runs the same integer arithmetic, so the exported
witness matches the field circuit bit-for-bit.

\paragraph{Transformer block.}
A pre-norm block~\cite{vaswani2017attention} computes
$\bX_{\mathrm{mid}}\!=\!\bX_{\mathrm{in}}+\Att(\LN(\bX_{\mathrm{in}}))$ and
$\bX_{\mathrm{out}}\!=\!\bX_{\mathrm{mid}}+\mathrm{FFN}(\LN(\bX_{\mathrm{mid}}))$,
with softmax attention $\softmax(\bQ\bK^\top\!/\sqrt d)\bV$ and
LayerNorm~\cite{ba2016layer}
$\bgamma\!\odot\!(\bx-\mu)/\sqrt{\sigma^2+\epsilon}+\bbeta$. The
non-arithmetic primitives (exp, division, square root) dominate SNARK
complexity in prior work.

\paragraph{Sumcheck.}
The sumcheck protocol~\cite{setty2020spartan,thaler2022proofs} reduces a
claim $H\!=\!\sum_{x\in\{0,1\}^n}\!f(x)$ for an MLE $f$ of individual
degree $\delta$ to a single evaluation $f(r)$ at random $r$, with
soundness error $\le n\delta/|\FF|$ and verifier cost $O(n\delta)$.
\sys{} uses the degree-2 variant for products of two MLEs, the cubic
variant for LayerNorm variance and our zerocheck-folded range proof,
and a multi-batched variant $\sum_k\lambda_k\sum_x f_k(x)g_k(x)$ for
cross-block batching.

\paragraph{Hyrax PCS.}
Hyrax~\cite{wahby2018doubly} arranges $2^n\!=\!2^{\nu+\sigma}$ MLE
evaluations as a $2^\nu\!\times\!2^\sigma$ matrix and commits to each
row via a vector Pedersen commitment over transparent generators. The
scheme is computationally binding under DL on $\GG$, requires no trusted
setup, and gives $O(\sqrt N)$ commitments, openings, and proof size.
Crucially, the row-Pedersen structure is linear:
$\Com(\bA)\!+\!\Com(\bB)\!=\!\Com(\bA\!+\!\bB)$, which \sys{} exploits to
verify residual connections without any proof.

\paragraph{Lasso.}
\label{ssec:lasso}
Lasso~\cite{setty2024unlocking,arun2024jolt,fomenko2025understanding}
proves that $N$ queries are consistent with a public sub-table
$T\!\in\!\FF^{2^m}$ via one batched sumcheck. A composite table that
decomposes \emph{additively} into $c$ sub-tables (Lasso's structured
condition) is committed as $c$ small tables of size $2^{B/c}$. Lasso's
defining property for our design is that prover and verifier cost
depend on table size and query count, but are independent of table
\emph{contents}: training the entries of $T$ has zero proving-cost
impact.

%-------------------------------------------------------------------------------
\section{The \sys{} Architecture}
\label{sec:arch}
%-------------------------------------------------------------------------------

\begin{figure*}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    font=\small,
    box/.style={rectangle, draw, rounded corners=2pt, minimum width=1.4cm,
                minimum height=0.55cm, align=center, fill=blue!4},
    op/.style={rectangle, draw, rounded corners=2pt, minimum width=1.4cm,
               minimum height=0.55cm, align=center, fill=green!8},
    ln/.style={rectangle, draw, rounded corners=2pt, minimum width=1.2cm,
               minimum height=0.55cm, align=center, fill=orange!10},
    add/.style={circle, draw, inner sep=0.5pt, font=\footnotesize, fill=white},
    arrow/.style={-Latex, thick},
    label/.style={font=\scriptsize, color=gray!70!black}
  ]
  % ----- Top row: attention sub-block -----
  \node[box] (xin) at (0,0) {$\bX_{\mathrm{in}}$};
  \node[ln,  right=0.4cm of xin]   (ln1)  {$\LN_1$};
  \node[op,  right=0.4cm of ln1]   (qkv)  {$\bW_{Q,K,V}$\\(ternary)};
  \node[ln,  right=0.4cm of qkv]   (qkn)  {$\LN_{Q}\!/\LN_{K}$\\(QK-norm)};
  \node[op,  right=0.4cm of qkn]   (lla)  {LLA\\(Lasso $\bphi$\\$+\,Z$ norm)};
  \node[op,  right=0.4cm of lla]   (woo)  {$\bW_O$\\(ternary)};
  \node[ln,  right=0.4cm of woo]   (aon)  {$\LN_{\mathrm{aon}}$\\(sandwich)};
  \node[add, right=0.35cm of aon]  (res1) {$+$};
  \draw[arrow] (xin.east) -- (ln1.west);
  \draw[arrow] (ln1.east) -- (qkv.west);
  \draw[arrow] (qkv.east) -- (qkn.west);
  \draw[arrow] (qkn.east) -- (lla.west);
  \draw[arrow] (lla.east) -- (woo.west);
  \draw[arrow] (woo.east) -- (aon.west);
  \draw[arrow] (aon.east) -- (res1.west);
  \draw[arrow] (xin.north) |- ($(xin.north)+(0,0.7)$) -| (res1.north);

  % ----- Bottom row: FFN sub-block -----
  \node[box,  below=1.3cm of xin]    (xmid) {$\bX_{\mathrm{mid}}$};
  \node[ln,   right=0.4cm of xmid]   (ln2)  {$\LN_2$};
  \node[op,   right=0.4cm of ln2]    (w1)   {$\bW_1$\\(ternary)};
  \node[op,   right=0.4cm of w1]     (phi)  {$\bphi$\\(Lasso)};
  \node[op,   right=0.4cm of phi]    (w2)   {$\bW_2$\\(ternary)};
  \node[add,  right=0.35cm of w2]    (res2) {$+$};
  \node[box,  right=0.4cm of res2]   (xout) {$\bX_{\mathrm{out}}$};
  \draw[arrow] (xmid.east) -- (ln2.west);
  \draw[arrow] (ln2.east)  -- (w1.west);
  \draw[arrow] (w1.east)   -- (phi.west);
  \draw[arrow] (phi.east)  -- (w2.west);
  \draw[arrow] (w2.east)   -- (res2.west);
  \draw[arrow] (res2.east) -- (xout.west);
  \draw[arrow] (xmid.north) |- ($(xmid.north)+(0,0.7)$) -| (res2.north);
  \draw[arrow] (res1.south) |- ($(res1.south)+(0,-0.45)$) -| (xmid.north);

  \node[draw, dashed, rounded corners,
        fit=(ln1)(aon)(res1)(xmid)(xout)(res2),
        inner sep=8pt, label={[label]above:Single \sys{} block (one of $L$);\;5 LayerNorms (orange)}] {};
  \node[font=\scriptsize, color=gray!70!black, below=2.0cm of xmid,
        align=left, text width=0.95\textwidth] (note) {%
    \textbf{Cross-block batching.} The five LNs per block (LN$_1$,
    QK-norm $\LN_Q\!/\LN_K$, sandwich $\LN_{\mathrm{aon}}$, LN$_2$),
    plus the final LN, are folded into one
    \emph{model-level} batched LN proof. Ternary projections fold into
    \texttt{qkv}/\texttt{oproj}/\texttt{ffn} sumchecks; LLA inner products
    fold into \texttt{batch\_attn\_out}/\texttt{ctx}; the row-normaliser $Z$
    is enforced by a cubic multi-batched sumcheck with range proofs on the
    floor-division residuals (\S\ref{ssec:LLA}); Lasso $\bphi$ folds into
    one global multi-instance proof per lookup family; range checks fold
    into one shared multiplicity commitment per width bucket;
    residuals are homomorphic Hyrax addition.
  };
\end{tikzpicture}}
\caption{A single \sys{} block. Boxes are matrices; rounded green nodes are
operations the SNARK proves; orange nodes are LayerNorms. \emph{Sandwich
norm} ($\LN_{\mathrm{aon}}$) is applied to the attention output before the
residual; \emph{QK-norm} ($\LN_Q$, $\LN_K$) is applied to $\bQ,\bK$ after the
ternary projection and before $\bphi$. Ternary projections contain no field
multiplications in the contraction; Lasso lookup cost is independent of
table contents; residual $+$ is checked by homomorphic addition of Hyrax
commitments and requires no proof.}
\label{fig:block}
\end{figure*}

A \sys{} block (Figure~\ref{fig:block}) is a sandwich-normalised
linear-attention block. With $\bX_{\mathrm{in}}\!\in\!\FF^{T\times d}$,
the block computes
\begin{align*}
  \bX_{\mathrm{n1}}      &= \LN_1(\bX_{\mathrm{in}}),\\
  \bQ,\bK,\bV           &= \bX_{\mathrm{n1}}\bW_{Q,K,V} \;\;\text{(ternary)},\\
  \bQ^{\mathrm{n}},\bK^{\mathrm{n}} &= \LN_Q(\bQ),\;\LN_K(\bK),\\
  \bN                   &= \bphi(\bQ^{\mathrm{n}})\bigl(\bphi(\bK^{\mathrm{n}})^\top\bV\bigr),\\
  Z_t                   &= \bphi(\bq_t^{\mathrm{n}})^{\!\top}\!{\textstyle\sum_s}\bphi(\bk_s^{\mathrm{n}}),\\
  \bY^{\mathrm{att}}_{tj}&= \lfloor S\!\cdot\!\bN_{tj}/Z_t\rfloor,\\
  \bX_{\mathrm{mid}}    &= \bX_{\mathrm{in}} + \LN_{\mathrm{aon}}(\bY^{\mathrm{att}}\bW_O+\bb_O),\\
  \bX_{\mathrm{out}}    &= \bX_{\mathrm{mid}} + \bphi(\LN_2(\bX_{\mathrm{mid}})\bW_1)\bW_2.
\end{align*}
After $L$ such blocks the model applies a final LayerNorm and a
ternary LM head; the model contains $5L\!+\!1$ LayerNorms in total.
The scale $S$ re-introduces fractional precision before the integer
floor division.

Two of the five LayerNorms per block are not in a textbook
transformer but are essential for our integer pipeline.
\emph{QK-norm} ($\LN_Q,\LN_K$) pins the range of $\bQ,\bK$ so that
$\bphi(\bQ^{\mathrm n}),\bphi(\bK^{\mathrm n})$ and $Z$ stay in the
lookup-table range under any per-layer ternary scale.
\emph{Sandwich norm} ($\LN_{\mathrm{aon}}$) absorbs the unbounded
magnitude of linear attention before the residual: softmax is
row-stochastic and naturally bounded, but
$\bphi(\bQ)(\bphi(\bK)^\top\bV)$ is not, and an unnormalised residual
destabilises downstream LayerNorms. Both share one LN sub-protocol
(\S\ref{ssec:ln}); their only cost is extra LN witnesses, folded into
the model-level batched LN proof and the bucketed range batch.

\subsection{Learnable Lookup Attention (LLA)}
\label{ssec:LLA}

Standard softmax attention requires $T^2$ exponentials and $T$ row-wise
divisions. We replace the kernel with a positive-definite inner product
$\kappa(\bq,\bk)=\langle\bphi(\bq),\bphi(\bk)\rangle$ on the post-QK-norm
tensors $\bQ^{\mathrm{n}},\bK^{\mathrm{n}}$, where
$\bphi:\RR^{d_h}\!\to\!\RR^{d_h}$ is applied element-wise and realised as a
structured lookup (\S\ref{ssec:phi}). The block computes the un-normalised
numerator $\bN$ and the per-row denominator $Z$ as
\begin{equation}
  \bN_{tj} = \bphi(\bq_t^{\mathrm{n}})^{\!\top}\!\bigl({\textstyle\sum_s}\bphi(\bk_s^{\mathrm{n}})\bv_s^{\!\top}\bigr)_j,\;\;
  Z_t = \bphi(\bq_t^{\mathrm{n}})^{\!\top}\!{\textstyle\sum_s}\bphi(\bk_s^{\mathrm{n}}).
\end{equation}
The context matrix $\bM=\sum_s\bphi(\bk_s^{\mathrm{n}})\bv_s^{\!\top}\!\in\!\FF^{d_h\times d_h}$
depends only on the keys/values and is computed once: per-token output
cost is $O(d_h^2)$ and the per-block prover cost becomes $O(T d_h^2)$
instead of $O(T^2 d_h)$~\cite{katharopoulos2020transformers}.

\paragraph{Floor-division proof for row normalisation.}
Unlike the previous prototype, \sys{} now proves the row normalisation
$\bY^{\mathrm{att}}_{tj}=\lfloor S\!\cdot\!\bN_{tj}/Z_t\rfloor$ end-to-end.
The witness includes auxiliary tensors $\bN,\bY^{\mathrm{att}},Z,\mathrm{rem},\mathrm{diff}$
where $\mathrm{rem}_{tj}=S\bN_{tj}-Z_t\bY^{\mathrm{att}}_{tj}$ and
$\mathrm{diff}_{tj}=Z_t-1-\mathrm{rem}_{tj}$, all committed via Hyrax. At a
single random point $r_{\mathrm{norm}}\!\in\!\FF^{t_{\mathrm{bits}}+d_{\mathrm{bits}}}$
the prover runs one cubic multi-batched sumcheck enforcing
\begin{align}
  S\!\cdot\!\bN \;-\; \mathrm{rem} \;-\; Z\!\cdot\!\bY^{\mathrm{att}} &\;=\; 0, \\
  \lambda\bigl(\,Z \;-\; 1 \;-\; \mathrm{rem} \;-\; \mathrm{diff}\,\bigr) &\;=\; 0,
\end{align}
merged into one cubic claim by the Fiat--Shamir scalar $\lambda$. Range
proofs on $\mathrm{rem},\mathrm{diff}$ (added to the 64-bit width bucket of
\S\ref{ssec:range}) prove $\mathrm{rem},\mathrm{diff}\!\geq\!0$, which
combined with the second identity gives $0\!\leq\!\mathrm{rem}\!<\!Z$~--- the
floor-division witness. A second sumcheck binds $Z_t$ to
$\bphi(\bq_t^{\mathrm{n}})\!\cdot\!\sum_s\bphi(\bk_s^{\mathrm{n}})$ via the
already-committed $\bphi(\bQ^{\mathrm{n}}),\bphi(\bK^{\mathrm{n}})$ MLEs; a
causal variant binds $Z$ to the prefix sum.

\paragraph{Lookup binding to post-QK-norm tensors.}
The Lasso queries for $\bphi$ are evaluated on the post-QK-norm tensors,
not on the pre-norm $\bQ,\bK$. The prover commits both
$C_{\bQ},C_{\bK}$ (used by the QKV projection sumcheck) \emph{and}
$C_{\bQ^{\mathrm{n}}},C_{\bK^{\mathrm{n}}}$ (used as the $\bphi$ inputs);
the model-level $\bphi(\bQ),\bphi(\bK)$ Lasso binds its query indices to
$C_{\bQ^{\mathrm{n}}},C_{\bK^{\mathrm{n}}}$ via one shared Hyrax batch open,
and the QKV cross-block sumcheck binds $C_{\bQ},C_{\bK}$ separately. The
LayerNorm sub-proof for $\LN_Q$/$\LN_K$ (folded into the model-level
batched LN proof) ties the two together.

\subsection{Additive sub-table feature map}
\label{ssec:phi}

Rather than committing to a fixed kernel, \sys{} parametrises
$\bphi$ as a sum of \emph{learnable} sub-table lookups. Let $\bits$
be the input bit-width and $\chunks\!\mid\!\bits$ the decomposition
depth, $\chunkbits\!=\!\bits/\chunks$ bits per chunk. For scalar $x$,
define the integer index
$x_{\mathrm{int}}=\mathrm{clamp}(\lfloor x/s\rfloor,0,2^\bits-1)$ and
chunk extractor
$\chi_i(x_{\mathrm{int}})=(x_{\mathrm{int}}\!\gg\!i\chunkbits)\,\&\,(2^{\chunkbits}\!-\!1)$.
Then
\begin{equation}
  \bphi(x) \;=\; \sum_{i=0}^{\chunks-1} T_i\bigl[\chi_i(x_{\mathrm{int}})\bigr],
  \label{eq:phi}
\end{equation}
where $T_0,\dots,T_{\chunks-1}\!\in\!\RR^{2^{\chunkbits}}$ are
learnable sub-tables, initialised so $\sum_i T_i\!\approx\!\GELU$ and
trained end-to-end with the straight-through
estimator~\cite{bengio2013estimating}. For $\bits\!=\!16,\chunks\!=\!2$ this
replaces a $65{,}536$-entry monolithic table with two $256$-entry
tables: \emph{exactly} the Lasso structured condition.

The crucial property exploited by \sys{} is that, under
Lasso (\S\ref{ssec:lasso}), the cost of proving a batch of lookups
against $T_i$ depends only on the size of $T_i$ and on the number of
queries; it does not depend on the values of $T_i$. The model can
therefore train the entries of every $T_i$ from scratch without
incurring any additional proving cost. We implement
\texttt{StructuredLookupActivation} on the Python side with a
sub-table forward pass and an STE backward pass; the integer indices
emitted by the Python forward are bit-identical to those proven on
the Rust side.

\subsection{Ternary projections}
\label{ssec:ternary}
All projection matrices $\bW_Q,\bW_K,\bW_V,\bW_O,\bW_1,\bW_2$ and the
LM head are constrained to entries in $\calT=\{-1,0,+1\}$ scaled by
a single learnable scalar $\alpha\!\in\!\FF$ per layer, following
BitNet b1.58~\cite{ma2025bitnet}. The forward pass is
$\by=\alpha(\bx\bW)+\bb$ with $\bW\!\in\!\calT^{m\times n}$. Training
uses the magnitude-thresholded mapping
$w\!\mapsto\!\mathrm{sign}(w)\!\cdot\![|w|\!>\!0.7\overline{|w|}]$ with
STE; the scalar $\alpha$ is multiplied in \emph{after} the matrix
product so the in-circuit weight stays ternary.

In the SNARK, a degree-2 sumcheck for a ternary contraction
\[
  \bY(r_t,r_d)-\bb(r_d)=\alpha\!\sum_k\bX(r_t,k)\bW(k,r_d)
\]
materialises only the cheap vector $\alpha\!\cdot\!\bW(\cdot,r_d)$,
never a field weight matrix; the prover folds the equality polynomial
against ternary entries with $+\!=$/$-\!=$/skip rather than field
multiplications. The Rust prover represents weights as
\texttt{Vec<Vec<TernaryValue>>} with
\texttt{TernaryValue}$\,\in\,$\texttt{\{ONE, ZERO, MINUSONE\}} and
folds via a specialised \texttt{eval\_cols\_ternary} helper. A
self-contained Lasso check
(\texttt{attention/ternary\_check.rs}) over the 4-element table
$[0,1,r-1,0]$ enforces ternarity by an integer encoding
$\mathsf{enc}(w)\!\in\!\{0,1,2\}$; this module is currently a
stand-alone protocol and is a candidate to fold into setup-time
preprocessing so the verifying key carries a binding proof that
$\bW\!\in\!\calT^{m\times n}$.

\subsection{Constraint-fused LayerNorm}
\label{ssec:ln}
Let $\bx\!\in\!\ZZ^d$ be a row,
$\mathrm{sum\_x}\!=\!\sum_j x_j$,
$\mathrm{sq\_sum\_x}\!=\!\sum_j x_j^2$,
$\mathrm{sum\_x\_sq}\!=\!\mathrm{sum\_x}^2$, and
$\sigma\mathrm{\_sq\_scaled}\!=\!(d\sigma)^2$ four auxiliary witness
fields with $\sigma\!=\!\lfloor\sqrt{\mathrm{var\_x}}/d\rfloor$. The integer
variance is recovered from
$\mathrm{var\_x}\!=\!d\bigl(d\,\mathrm{sq\_sum\_x}-\mathrm{sum\_x\_sq}\bigr)$,
which expands $\sum_j(d x_j\!-\!\mathrm{sum\_x})^2$ without ever taking
$\mathrm{sum\_x}^2$ outside the sumcheck. \sys{} proves the entire LayerNorm
in five sub-protocols with no field divisions and no square roots:

\begin{enumerate}
  \item \textbf{Mean (degree-2).} Sumcheck for
        $\widetilde{\mathrm{sum\_x}}(r_t)\!=\!\sum_j x_{\mathrm{col}}(j)$
        where $x_{\mathrm{col}}\!=\!\widetilde\bx.\mathrm{fix\_row}(r_t)$.
  \item \textbf{Square-sum (cubic).} Sumcheck for
        $\widetilde{\mathrm{sq\_sum\_x}}(r_t)\!=\!\sum_j x_{\mathrm{col}}(j)^3$
        via three copies of $x_{\mathrm{col}}$ (the third factor folds
        $1\!=\!x/x$ so the round polynomial stays multilinear in degree
        three).
  \item \textbf{Sigma-residual (cubic multi-batched).} Bind
        $\widetilde{\mathrm{sum\_x\_sq}}(r_{\sigma_t})\!=\!\sum_i\eq(r_{\sigma_t},i)\widetilde{\mathrm{sum\_x}}(i)^2$
        and
        $\widetilde{\sigma\mathrm{\_sq}}(r_{\sigma_t})\!=\!\sum_i\eq(r_{\sigma_t},i)(d\widetilde\sigma(i))^2$
        in a single cubic multi-batched sumcheck under a Fiat--Shamir scalar
        $\lambda_{\mathrm{sig}}$; the verifier then reconstructs
        $\mathrm{var\_x}(r_{\sigma_t})$ algebraically.
  \item \textbf{$\sigma$ floor-sqrt range proof.} Show $\sigma$ is the
        integer floor square-root by range-checking
        $\mathrm{var\_x}\!-\!(d\sigma)^2\!\geq\!0$ and
        $(d\sigma\!+\!d)^2\!-\!1\!-\!\mathrm{var\_x}\!\geq\!0$ through the
        chunked Lasso range proof of \S\ref{ssec:range}.
  \item \textbf{Y constraint fusion.} For a single random
        $(r_{y_t},r_{y_d})$, audit
        $\bgamma\!\odot\!(d\bx-\mathrm{sum\_x})+\bbeta\cdot d\sigma\!\approx\!d\sigma\!\cdot\!\by$
        (up to integer rounding) by another range proof on the
        $\mathrm{lo}/\mathrm{hi}$ residuals; the $\gamma\bX$ and
        $\sigma\bY$ legs are fused into a single multi-batched cubic
        sumcheck.
\end{enumerate}
The prover commits three internal Hyrax commitments per LN
($\mathrm{sum\_x}$, $\mathrm{sq\_sum\_x}$, $\sigma$);
$\mathrm{sum\_x\_sq}$ and $\sigma\mathrm{\_sq\_scaled}$ are bound
algebraically by step~3 and need no separate commitment. The verifier
evaluates $\bgamma$ and $\bbeta$ directly from the VK in $O(d)$ field
operations per layer; everything else is one shared opening per witness.

\paragraph{Model-level batched LN proof.}
The end-to-end prover does not emit one independent LN sub-proof per LN.
Instead, after Phase~1 has absorbed all $5L+1$ LN IO commitments
($\LN_1$, $\LN_Q$, $\LN_K$, $\LN_{\mathrm{aon}}$, $\LN_2$ per block, plus
the final LN), one call to \textsc{ProveLNsBatched} folds the mean,
variance, $\sigma$-floor, and Y-fusion sub-claims of every LN into a single
batched cubic sumcheck under Fiat--Shamir powers of an independent
challenge. Soundness follows from the same Schwartz--Zippel argument as
cross-block batching (Theorem~\ref{thm:cross-block}) with the role of $L$
played by $5L+1$.

\subsection{Globally batched range proof}
\label{ssec:range}
The range proof shows $V[i]\!\in\![0,2^{\mathrm{bits}})$ via a chunked Lasso
approach over identity tables, with $\mathrm{CHUNK\_BITS}\!=\!16$. The
implementation supports two width buckets, $\mathrm{bits}\!\in\!\{32,64\}$,
and routes each witness to the narrowest bucket that contains it. The
LayerNorm $\sigma$/$y$ witnesses use 64-bit range proofs
($\mathrm{LAYERNORM\_RANGE\_BITS}\!=\!64$); the optional attention-
normalisation residuals $\mathrm{rem},\mathrm{diff}$ (\S\ref{ssec:LLA}) are
also 64-bit. Each non-empty bucket is processed independently and produces
one $m_{\mathrm{com}}$.

For $B$ witnesses in a single bucket with $C\!=\!\lceil\mathrm{bits}/16\rceil$
chunks each (so $C\!\in\!\{2,4\}$), Phase~1 commits all
$V^{(b)}_0,\dots,V^{(b)}_{C-1}$ via Hyrax and merges all chunk arrays into
one global multiplicity vector
$m[v]=\sum_{b,c,i}[\mathrm{chunk}^{(b,c)}[i]\!=\!v]$ committed once.
Phase~2 runs one degree-2 sumcheck per witness binding $V^{(b)}$ to a
random $r_v^{(b)}$ and checks
$V^{(b)}(r_v^{(b)})=\sum_c 2^{16c}V^{(b)}_c(r_v^{(b)})$ through one Hyrax
batch open per witness. Phase~3 opens $m(r_m)$ once and checks the LogUp
identity~\cite{habo2024logup} against the identity table
$T_{\mathrm{id}}[i]\!=\!i$. The transcript-ordering invariant
(Theorem~\ref{thm:range}) is that $m_{\mathrm{com}}$ is absorbed before
any Phase~2 sumcheck challenge is sampled.

\paragraph{Zerocheck-folded inverse polynomial.}
The LogUp identity requires per-chunk inverse polynomials
$h_{c}[i]\!=\!1/(\alpha-V_{c}[i])$ for a Fiat--Shamir challenge $\alpha$.
A na\"ive implementation Hyrax-commits each $h_{c}$ and opens it
alongside the chunk commitments, costing $B\!\cdot\!C$ extra Pedersen
MSMs per bucket. We avoid those commitments by folding a zerocheck of
\[
  h(x)\cdot(\alpha-V(x))\,-\,1\;\equiv\;0\qquad(x\!\in\!\{0,1\}^{n})
\]
into the bucket's existing cubic sumcheck. Concretely, after the
bucket claim
$\Lambda\!=\!\sum_p w_p\!\cdot\!\bigl(\sum_i h_p[i]+\beta\!\cdot\!n_{\mathrm{pad}}\bigr)$
is absorbed, the verifier samples $\gamma_{\mathrm{fold}}\!\in\!\mathbb{F}$
and $r_{zc}\!\in\!\mathbb{F}^{n}$, and the prover proves
\[
  \sum_{x}\Bigl[\;
     w(x)\!\cdot\!h(x)\!\cdot\!\bigl(1+\beta(\alpha-V(x))\bigr)
     \;+\;\gamma_{\mathrm{fold}}\!\cdot\!\mathrm{eq}(r_{zc},x)\!\cdot\!\bigl(h(x)(\alpha-V(x))-1\bigr)
  \Bigr]\;=\;\Lambda.
\]
The first term is the standard LogUp bucket claim; the second is the
zerocheck term, which sums to zero on the boolean hypercube iff
$h\!=\!1/(\alpha-V)$ pointwise. The integrand remains cubic in
$(h,V,w,\mathrm{eq})$, so per-round prover cost is unchanged. We
pre-fill $h$'s pair-padding slots with $\alpha^{-1}$ so that the
zerocheck term vanishes there ($\alpha^{-1}\!\cdot\!\alpha-1\!=\!0$),
keeping the bucket claim well-formed without an indicator mask.

Crucially, $\Lambda$ is absorbed into the transcript \emph{before}
$\gamma_{\mathrm{fold}}$ and $r_{zc}$ are sampled. The terminal
$h(r_{\mathrm{full}})$ is sent as a scalar (not opened against any
commitment) and is bound only by the sumcheck's round-polynomial
consistency together with the combined-identity check; the chunk
evaluation $V(r_{\mathrm{full}})$ remains bound to the Hyrax-committed
$V_c$ via the per-chunk openings at the bucket challenge $r_k$.

The model-level prover collects \emph{all} range witnesses across the
entire model into a single bucket-routed list: $10L+2$ LN witnesses
($\sigma,y$ for the five LNs per block plus the final LN's two), and
$2L$ extra witnesses ($\mathrm{rem},\mathrm{diff}$ per block) when
normalised attention is enabled. Sharing one $m_{\mathrm{com}}$ per
bucket therefore saves $(B\!-\!1)\!\times\!\sqrt{2^{16}}$ G1 MSMs per
bucket, and the savings scale with $L$ rather than being capped at the
per-block level. The zerocheck fold further eliminates $B\!\cdot\!C$
Hyrax commitments and as many Hyrax openings per bucket compared with
the na\"ive LogUp scheme.

%-------------------------------------------------------------------------------
\section{The \sys{} Protocol}
\label{sec:protocol}
%-------------------------------------------------------------------------------

This section gives the protocol. Algorithm~\ref{alg:setup} describes
preprocessing; Algorithm~\ref{alg:prove} the model-level prover;
Algorithm~\ref{alg:phase1} the per-block ``Phase~1'' that commits
intermediate matrices and proves both LayerNorms;
Algorithm~\ref{alg:cross-block} the cross-block batched sumcheck
primitive; and Algorithm~\ref{alg:verify} the verifier. The LayerNorm
sub-prover (Algorithm~\ref{alg:layernorm}) and the globally batched
range proof are described in \S\ref{ssec:ln} and \S\ref{ssec:range}.

\subsection{Preprocessing}

\begin{algorithm}[H]
\caption{$\sys{}.\Setup(\theta;\,\text{params})$}
\label{alg:setup}
\begin{algorithmic}[1]
\Require model weights
$\theta\!=\!\{\bW_Q^\ell,\bW_K^\ell,\bW_V^\ell,\bW_O^\ell,\bW_1^\ell,\bW_2^\ell,\bgamma^\ell,\bbeta^\ell\}_{\ell=1}^L$,
final-LN params, LM head $\bW_h$;
params $(T,d,d_{ff},d_h,\bits,\chunks)$.
\Ensure proving key $\poppk$, verifying key $\popvk$.
\State derive Hyrax generators $g_j\!\gets\!\mathrm{SHA3}(\texttt{"piformer-hyrax-gen"}\!\Vert\! j)\cdot G$.
\For{each weight matrix $\bW\!\in\!\theta\cup\{\bW_h\}$}
  \State $C_\bW\gets\Com(\bW)$ \Comment{Hyrax commit to ternary entries}
\EndFor
\For{each lookup sub-table $T_0,\dots,T_{\chunks-1}$}
  \State $C_{T_i}\gets\Com(T_i)$
\EndFor
\State $\popvk\gets\{C_\bW\}\cup\{C_{T_i}\}\cup\{\bgamma^\ell,\bbeta^\ell\}\cup\{g_j\}$
\State $\poppk\gets\popvk\cup\{\text{raw ternary arrays, sub-tables}\}$
\State \Return $(\poppk,\popvk)$
\end{algorithmic}
\end{algorithm}

The verifying key contains \emph{no} raw weights, only commitments.
The proving key additionally carries the ternary $\{-1,0,+1\}$
arrays and the lookup sub-tables so the prover can fold the equality
polynomial against them in $O(Td)$ time without any field
multiplications.

\subsection{The model-level prover}

\begin{algorithm}[H]
\caption{$\sys{}.\Prove(\poppk,W,\mathsf{inst})$}
\label{alg:prove}
\begin{algorithmic}[1]
\Require proving key $\poppk$, witness $W$, instance $\mathsf{inst}$
(Lasso queries for $\bphi(\bQ),\bphi(\bK),\bphi(\bM)$ at every block).
\Ensure model proof $\piproof$.

\State $\mathsf{FS}.\mathsf{init}(\texttt{"piformer"})$;\;
       $C_{\bx_{\mathrm{in}}}\!\gets\!\Com(\bx_{\mathrm{in}})$;\;
       $\mathsf{FS}.\mathsf{absorb}(C_{\bx_{\mathrm{in}}})$
\Phase{1: per-block commit $+$ LN IO absorption}
\State $C_{\mathrm{cur}}\!\gets\!C_{\bx_{\mathrm{in}}}$
\For{$\ell\!=\!1,\dots,L$}
  \State $p_\ell\!\gets\!\textsc{Phase1}(W_\ell,C_{\mathrm{cur}},\poppk_\ell,\mathsf{FS})$
  \Comment{commits 5 LN IO matrices, $\bphi$ inputs, optional attn-norm aux}
  \State $C_{\bX_{\mathrm{mid}}^\ell}\!\gets\!C_{\mathrm{cur}}+p_\ell.C_{\LN_{\mathrm{aon}}.y}$
         \Comment{sandwich-norm residual}
  \State $C_{\mathrm{cur}}\!\gets\!C_{\bX_{\mathrm{mid}}^\ell}+p_\ell.C_{\bO_{\mathrm{ffn}}}$
\EndFor
\Phase{2: global random point}
\State $r_{td}\!=\!(r_t\!\Vert\!r_{\mathrm{out}})\gets\!\mathsf{FS}.\mathsf{challenge}(t_{\mathrm{bits}}+d_{\mathrm{bits}})$
\Phase{3: four cross-block sumchecks}
\State $\pi_{qkv}\!\gets\!\textsc{CrossBatch}(\{f^{qkv}_\ell,g^{qkv}_\ell\}_\ell,\mathsf{FS})$
\State $\pi_{op}\!\gets\!\textsc{CrossBatch}(\{f^{op}_\ell,g^{op}_\ell\}_\ell,\mathsf{FS})$
\State $\pi_{ao}\!\gets\!\textsc{CrossBatch}(\{f^{ao}_\ell,g^{ao}_\ell\}_\ell,\mathsf{FS})$
\State $\pi_{ac}\!\gets\!\textsc{CrossBatch}(\{f^{ac}_\ell,g^{ac}_\ell\}_\ell,\mathsf{FS})$
\Phase{4: attention normalisation sumcheck (when normalised mode)}
\State $r_{\mathrm{norm}}\!\gets\!\mathsf{FS}.\mathsf{challenge}(t_{\mathrm{bits}}+d_{\mathrm{bits}})$;\;
       $\lambda\!\gets\!\mathsf{FS}.\mathsf{challenge}$
\State $\pi_{\mathrm{atn}}\!\gets\!\textsc{ProveCubicMulti}\bigl(
        \{\bN_\ell,Z_\ell,\bY^{\mathrm{att}}_\ell,\mathrm{rem}_\ell,\mathrm{diff}_\ell\}_\ell,\,\lambda,\,\mathsf{FS}\bigr)$
\State $\pi_{Z}\!\gets\!\textsc{CrossBatch}(\{\bphi(\bQ^{\mathrm{n}})_\ell,\sum_s\bphi(\bK^{\mathrm{n}})_\ell\}_\ell,\mathsf{FS})$
\Phase{5: global FFN Lasso $+$ FFN-Y/M sumchecks}
\State commit $\{\bA_\ell,\bM_\ell\}_\ell$;\;
       $\pi_{\mathrm{lasso}}^{\mathrm{ffn}}\!\gets\!\Prove\text{-}\mathrm{LassoMulti}\bigl(\{\mathsf{inst}_\ell.\bphi(\bM)\}_\ell,\{(C_{\bA_\ell},\bA_\ell)\}_\ell,\mathsf{FS}\bigr)$
\State bind FFN indices to $\{C_{\bM_\ell}\}_\ell$ via one Hyrax batch open
\State $\pi_{ffy}\!\gets\!\textsc{CrossBatch}(\{f^{ffy}_\ell,g^{ffy}_\ell\}_\ell,\mathsf{FS})$
\State $\pi_{ffm}\!\gets\!\textsc{CrossBatch}(\{f^{ffm}_\ell,g^{ffm}_\ell\}_\ell,\mathsf{FS})$
\Phase{6: bucketed range batch + model-level batched LN proof}
\For{$\mathrm{bits}\!\in\!\{32,64\}$}
  \State $\pi_{\mathrm{rng},\mathrm{bits}}\!\gets\!\textsc{ProveRangeBatched}(\text{all witnesses in bucket},\mathrm{bits},\mathsf{FS})$
\EndFor
\State $\pi_{\mathrm{ln}}\!\gets\!\textsc{ProveLNsBatched}(\text{all $5L\!+\!1$ LNs},\mathsf{FS})$
\Phase{7: LM head $+$ $\mu$-challenges $+$ batch opens}
\State $\pi_{\mathrm{head}}\!\gets\!\Prove\text{-}\mathrm{Proj}(W.\text{LM-head},\mathsf{FS})$
\State drain 13 deferred-accumulator $\mu$-challenges plus 2 read-only $\rho$
fuse challenges; finalise via \texttt{finalize\_many\_with\_mus} (one MSM
pair per Hyrax-params group)
\State $\pi_{\mathrm{inter}}\!\gets\!\Open\text{-}\mathrm{Batch}\bigl(\{\bQ_\ell,\bK_\ell,\bV_\ell,\bO_{\mathrm{att}}^\ell,\bO_{\mathrm{ffn}}^\ell\}_{\ell=1}^L,\,r_{td}\bigr)$
\State $\{\pi_{\mathrm{open}}^{(j)}\}\!\gets\!\Open\text{-}\mathrm{Batch}(\{\bX_{\mathrm{n1}}^\ell,\bW_Q^\ell,\dots,C_{\bphi(\bQ^{\mathrm{n}})_\ell},C_{\bphi(\bK^{\mathrm{n}})_\ell}\}_\ell,\,r^{(j)})$
\State (when normalised) $\Open\text{-}\mathrm{Batch}([\bN\!\mid\!\bY^{\mathrm{att}}\!\mid\!\mathrm{rem}\!\mid\!\mathrm{diff}]_\ell,\,r_{\mathrm{norm}})$
\Phase{8: global $\bphi(\bQ^{\mathrm{n}}),\bphi(\bK^{\mathrm{n}})$ Lasso}
\State commit $\{\bphi(\bQ^{\mathrm{n}})_\ell,\bphi(\bK^{\mathrm{n}})_\ell\}_\ell$;\;
       bind indices to $\{C_{\bQ^{\mathrm{n}}_\ell},C_{\bK^{\mathrm{n}}_\ell}\}$
\State $\pi_{\mathrm{lasso}}^{\mathrm{qk}}\!\gets\!\Prove\text{-}\mathrm{LassoMulti}\bigl(\{\mathsf{inst}_\ell.\bphi(\bQ^{\mathrm{n}}),\mathsf{inst}_\ell.\bphi(\bK^{\mathrm{n}})\}_\ell,\dots,\mathsf{FS}\bigr)$
\State \Return $\piproof$
\end{algorithmic}
\end{algorithm}

The key structural choices are:
\emph{(i) one global random point $r_{td}$ is drawn after all per-block
commitments are absorbed}, so all six cross-block sumchecks can share
it; \emph{(ii) roughly two dozen cross-cutting Hyrax accumulators}
(LayerNorm row / $td$, projection W/b, LM-head W/b, range $\sigma$/Y/$m$,
intermediate, quantization-FFN / -QK / -$m$, attn-norm, five cross-block
batch-open groups) are pushed by every claim and finalised once at the end
of the \emph{verifier}, with one MSM pair per group of shared Hyrax
parameters (MSM-fused across same-params groups via
\texttt{finalize\_many\_with\_mus}); \emph{(iii) residuals
are verified by linearity of the Hyrax PCS:}
$C_{\bX_{\mathrm{mid}}}=C_{\bX_{\mathrm{in}}}+C_{\bO_{\mathrm{att}}}$
is computed by the verifier from the existing commitments and
requires no proof.

\paragraph{Quantization-proof binding.}
\label{ssec:quant}
The Lasso protocol works over integer indices $\mathsf{idx}_j\!\in\![0,2^B)$,
but the committed input tensors $C_{\bM_\ell},C_{\bQ^{\mathrm{n}}_\ell},C_{\bK^{\mathrm{n}}_\ell}$
carry \emph{signed} field-element values. \sys{} bridges this gap with two
quantization sub-proofs, $\pi_{\mathrm{quant}}^{\mathrm{ffn}}$ and
$\pi_{\mathrm{quant}}^{\mathrm{qk}}$. Per layer the verifier knows the
quantization parameters $(s_{\mathrm{num}},s_{\mathrm{den}})$ from the VK
(with $s_{\mathrm{num}}$ a power of two) and the centred zero-point
$\mathsf{zp}=2^{B-1}$. The prover commits a remainder tensor
$\mathsf{rem}_j\!=\!r_j s_{\mathrm{den}}+\lfloor s_{\mathrm{num}}/2\rfloor
-s_{\mathrm{num}}(\mathsf{idx}_j-\mathsf{zp})$ via Hyrax, adds $\mathsf{rem}$
to the global range batch with bit-width $\log_2 s_{\mathrm{num}}$, and at a
Fiat--Shamir random point $\mathbf r$ opens both $\widetilde r$ and
$\widetilde{\mathsf{rem}}$ via batched Hyrax. The verifier evaluates the
\emph{public} index list $\widetilde{\mathsf{idx}}(\mathbf r)$ directly and
checks
\begin{equation}
  \widetilde r(\mathbf r)\,s_{\mathrm{den}} + \tfrac{s_{\mathrm{num}}}{2}\widetilde 1(\mathbf r)
  \;\stackrel{?}{=}\; s_{\mathrm{num}}\bigl(\widetilde{\mathsf{idx}}(\mathbf r)-\mathsf{zp}\,\widetilde 1(\mathbf r)\bigr) + \widetilde{\mathsf{rem}}(\mathbf r).
  \label{eq:quant}
\end{equation}
Combined with the remainder range proof, this argument pins every public
$\mathsf{idx}_j$ to the unique integer
$\lfloor (r_j s_{\mathrm{den}}+s_{\mathrm{num}}/2)/s_{\mathrm{num}}\rfloor+\mathsf{zp}$
consistent with the committed raw value $r_j$. A second, redundant
self-consistency check (the $\Open\text{-}\mathrm{Batch}$ over
\texttt{ffn\_lasso\_bind\_open} / \texttt{qk\_lasso\_bind\_open}) verifies
that the proof's index-MLE opening vector matches what the verifier
reconstructs from the public indices.

The two model-level Lasso families have a slight structural asymmetry.
The attention $\bphi(\bQ^{\mathrm n}),\bphi(\bK^{\mathrm n})$ proof is a
single committed-output \texttt{LassoMultiProof} (\texttt{all\_lasso\_proof})
that absorbs every $C_{\bphi(\bQ^{\mathrm n})_\ell},C_{\bphi(\bK^{\mathrm n})_\ell}$
before drawing batching challenges, then a second sumcheck inside the
multi-proof binds the combined lookup grand sum to the committed outputs.
The FFN family is rotated: a single cross-block $\pi_{ffy}$ degree-2
sumcheck reduces all $\bA_\ell$ leaves to one terminal claim
$\sum_\ell w_\ell\,\widetilde\bA_\ell(r_t,r_{k_{fy}})\!=\!c$, and a separate
\emph{Lasso terminal-eval proof}
($\pi_{ffn}^{\mathrm{term}}\!=\!\texttt{ffn\_a\_terminal\_proof}$) proves $c$
is the Lasso evaluation of the activation tables at the same terminal
point. This fuses the equivalent of a separate $C_{\bA_\ell}$ commitment,
a Lasso multi-proof output binding, and a Hyrax batch open of
$\{\bA_\ell\}$ into one sub-protocol with no proof-carried $\bA_\ell$.

\medskip

The two model-level Lasso instances merit separate mention. Each is a
\emph{committed-output} Lasso: before sampling the multi-Lasso batching
challenges, the prover absorbs the Hyrax commitments to the lookup
outputs ($C_{\bA_\ell}$ for FFN, $C_{\bphi(\bQ^{\mathrm{n}})_\ell}$ and
$C_{\bphi(\bK^{\mathrm{n}})_\ell}$ for attention). A second sumcheck inside
the Lasso multi-proof binds the combined lookup grand sum to the committed
output polynomials at the sumcheck terminal point. Lookup indices are
likewise bound to the committed \emph{post-norm} input tensors
($C_{\bM_\ell},C_{\bQ^{\mathrm{n}}_\ell},C_{\bK^{\mathrm{n}}_\ell}$) via
one shared Hyrax batch opening, so a cheating prover cannot pick query
indices independently of the committed inputs and cannot mix pre-norm and
post-norm tensors.

\subsection{Per-block Phase~1}

\begin{algorithm}[H]
\caption{$\textsc{Phase1}(W_\ell,C_{\mathrm{cur}},\poppk_\ell,\mathsf{FS})$}
\label{alg:phase1}
\begin{algorithmic}[1]
\Require block witness, current input commitment, block PK.
\Ensure intermediate commitments; \emph{no} LN sumcheck or range proof
emitted at this stage.
\State commit pre-/post-QK-norm $\bQ,\bK,\bV$:\;
       $C_\bQ,C_\bK,C_\bV,C_{\bQ^{\mathrm{n}}},C_{\bK^{\mathrm{n}}}$
\State commit LN outputs:\;
       $C_{\bX_{\mathrm{n1}}},C_{\LN_{\mathrm{aon}}.y},C_{\bX_{\mathrm{n2}}}$
\State commit O-projection / FFN outputs:\;
       $C_{\bO_{\mathrm{att}}},C_{\bO_{\mathrm{ffn}}}$
\State (when normalised attention) commit
       $C_{\bN},C_{\bY^{\mathrm{att}}},C_Z,C_{\mathrm{rem}},C_{\mathrm{diff}}$
\State absorb $\LN_1$ IO:\;
       $\mathsf{FS}.\mathsf{absorb}(C_{\mathrm{cur}},C_{\bX_{\mathrm{n1}}})$
\State absorb $C_\bQ,C_\bK,C_\bV$ and (when normalised) attn-norm aux commits
\State absorb $\LN_Q$ IO:\;
       $\mathsf{FS}.\mathsf{absorb}(C_\bQ,C_{\bQ^{\mathrm{n}}})$
\State absorb $\LN_K$ IO:\;
       $\mathsf{FS}.\mathsf{absorb}(C_\bK,C_{\bK^{\mathrm{n}}})$
\State $\mathsf{FS}.\mathsf{absorb}(C_{\bO_{\mathrm{att}}})$
\State absorb $\LN_{\mathrm{aon}}$ IO:\;
       $\mathsf{FS}.\mathsf{absorb}(C_{\bO_{\mathrm{att}}},C_{\LN_{\mathrm{aon}}.y})$
\State $C_{\bX_{\mathrm{mid}}^\ell}\!\gets\!C_{\mathrm{cur}}+C_{\LN_{\mathrm{aon}}.y}$
\Comment{sandwich-norm residual~1}
\State absorb $\LN_2$ IO:\;
       $\mathsf{FS}.\mathsf{absorb}(C_{\bX_{\mathrm{mid}}^\ell},C_{\bX_{\mathrm{n2}}})$
\State $\mathsf{FS}.\mathsf{absorb}(C_{\bO_{\mathrm{ffn}}})$
\State \Return all commitments
\end{algorithmic}
\end{algorithm}

The only truly per-block step is Phase~1: it commits the per-block
intermediate matrices and absorbs all five LN IO commitments (plus the
optional attention-normalisation aux commits) into the transcript in a
fixed order. \emph{No per-block LN sumcheck and no per-block range proof
are emitted at this stage}; both are deferred to model-level batched
calls (Phase~6 of Algorithm~\ref{alg:prove}). The remaining sub-protocols
(QKV, output projection, attention, FFN) are \emph{not} run independently
per block either; they are batched cross-block in Phases~3--5 of
Algorithm~\ref{alg:prove}. Each per-block proof carries the intermediate
Hyrax commitments, the FFN $C_{\bA_\ell},C_{\bM_\ell}$, the per-block
$C_{\bphi(\bQ^{\mathrm{n}})_\ell},C_{\bphi(\bK^{\mathrm{n}})_\ell}$, the
optional attention-normalisation aux commits, and the per-block scalar
evaluations consumed by the cross-block batch sumchecks.

\subsection{Cross-block batched sumcheck}

\begin{algorithm}[H]
\caption{$\textsc{CrossBatch}(\{f_\ell,g_\ell,H_\ell\}_{\ell=1}^L,\mathsf{FS})$}
\label{alg:cross-block}
\begin{algorithmic}[1]
\Require per-block MLEs $f_\ell,g_\ell\!:\!\FF^n\!\to\!\FF$ and per-block
targets $H_\ell$.
\Ensure batched proof $\piproof_{\mathrm{batch}}$ and shared random
point $r\!\in\!\FF^n$.
\For{$\ell\!=\!1,\dots,L$} absorb $H_\ell$ and per-block constants
into $\mathsf{FS}$ \EndFor
\State $\eta\!\gets\!\mathsf{FS}.\mathsf{challenge}(\texttt{"batch\_eta"})$
\State $H^\star\!\gets\!\sum_\ell \eta^{\ell-1}H_\ell$;\;
       $P(x)\!\gets\!\sum_\ell \eta^{\ell-1}f_\ell(x)g_\ell(x)$
\For{$i\!=\!1,\dots,n$}
  \State send $s_i(X)\!=\!\sum_{x\in\{0,1\}^{n-i}}P_i(X,x)$ at
         $X\!\in\!\{0,1,2\}$
  \State $r_i\!\gets\!\mathsf{FS}.\mathsf{challenge}$
\EndFor
\State \Return $(\eta,(s_i)_{i=1}^n)$,\,$r=(r_1,\dots,r_n)$
\end{algorithmic}
\end{algorithm}

The verifier recovers $H^\star$ from the per-block claims and $\eta$,
then runs the standard sumcheck verifier on
$\piproof_{\mathrm{batch}}$ to a final point $r$. From $r$ the
verifier checks the terminal claim
$P(r)=\sum_\ell \eta^{\ell-1}f_\ell(r)g_\ell(r)$ against the per-block
evaluations opened later via the cross-block batch opens. By
Theorem~\ref{thm:cross-block} this protocol replaces $L$ independent
sumcheck transcripts with one of length $n$ at a soundness cost of
an additive $(L-1)/|\FF|$.

\subsection{Constraint-fused LayerNorm prover}

The pseudocode below describes the LayerNorm protocol \emph{for one LN
witness}. The end-to-end prover does not invoke it once per LN; instead, it
collects all $5L+1$ LN witnesses (in fixed order: per block $\LN_1,\LN_Q,\LN_K,\LN_{\mathrm{aon}},\LN_2$;
then the final LN) and runs a single $\textsc{ProveLNsBatched}$ call as
Phase~6 of Algorithm~\ref{alg:prove}. That batched call shares the row
challenge $r_t$, the variance / Y-fusion sumchecks, and the deferred Hyrax
accumulator pushes across all LN instances, with per-LN claims combined by
Fiat--Shamir powers of an independent challenge.

\begin{algorithm}[H]
\caption{$\textsc{ProveLN}(W,\{C_\bx,C_\by\},(\mathsf{rps}_\sigma,\mathsf{rps}_y),\mathsf{FS})$}
\label{alg:layernorm}
\begin{algorithmic}[1]
\Require witness
$W=\{\bx,\by,\mathrm{sum\_x},\mathrm{sq\_sum\_x},\mathrm{sum\_x\_sq},
\sigma,\sigma\mathrm{\_sq\_scaled}\}$, where
$\mathrm{sum\_x\_sq}\!=\!\mathrm{sum\_x}^2$ and
$\sigma\mathrm{\_sq\_scaled}\!=\!(d\sigma)^2$ are square-witness fields
introduced so each algebraic identity collapses to a degree-${\le}3$
multilinear sumcheck.
\State commit $C_{\mathrm{sum\_x}},C_{\mathrm{sq\_sum\_x}},C_\sigma$;\;
       absorb IO and internals into $\mathsf{FS}$
\State $r_t\!\gets\!\mathsf{FS}.\mathsf{challenge}(t_{\mathrm{bits}})$
\State \emph{// 1. Mean sumcheck (degree-2)}
\State $(\pi^{\mathrm{mean}},r_d^{\mathrm{m}})\!\gets\!\Prove\text{-}\mathrm{Sumcheck}_2(\widetilde\bx_{\mathrm{col}},\mathbbm 1,\widetilde{\mathrm{sum\_x}}(r_t),\mathsf{FS})$
\State \emph{// 2. Square-sum sumcheck (cubic): $\widetilde{\mathrm{sq\_sum\_x}}(r_t)=\sum_j \widetilde\bx_{\mathrm{col}}(j)^3$ via three copies of $\widetilde\bx_{\mathrm{col}}$}
\State $(\pi^{\mathrm{sqs}},r_d^{\mathrm{v}})\!\gets\!\Prove\text{-}\mathrm{Sumcheck}_3(\widetilde\bx_{\mathrm{col}},\widetilde\bx_{\mathrm{col}},\widetilde\bx_{\mathrm{col}},\widetilde{\mathrm{sq\_sum\_x}}(r_t),\mathsf{FS})$
\State \emph{// 3. Sigma-residual sumcheck (cubic multi-batched, \texttt{sigma\_residual\_batch\_lambda})}
\State $r_{\sigma_t}\!\gets\!\mathsf{rps}_\sigma.r[..t_{\mathrm{bits}}]$;\;
       $\lambda_{\mathrm{sig}}\!\gets\!\mathsf{FS}.\mathsf{challenge}$
\State bind $\widetilde{\mathrm{sum\_x\_sq}}(r_{\sigma_t})=\sum_i\!\eq(r_{\sigma_t},i)\widetilde{\mathrm{sum\_x}}(i)^2$ and
       $\widetilde{\sigma\mathrm{\_sq}}(r_{\sigma_t})=\sum_i\!\eq(r_{\sigma_t},i)(d\widetilde\sigma(i))^2$
       in one cubic multi-batched sumcheck $\pi^{\mathrm{sig}}$
\State \emph{// 4. Sigma range proof (consumes $\mathsf{rps}_\sigma$)}
\State recover $\mathrm{var}(r_{\sigma_t})\!=\!d\bigl(d\widetilde{\mathrm{sq\_sum\_x}}(r_{\sigma_t})-\widetilde{\mathrm{sum\_x\_sq}}(r_{\sigma_t})\bigr)$;\;
       check $\mathrm{var}-(d\sigma)^2\!\geq\!0$ and $(d\sigma+d)^2-1-\mathrm{var}\!\geq\!0$ via $\mathsf{rps}_\sigma$
\State \emph{// 5. Y constraint fusion: $\gamma\bX$ + $\sigma\bY$ legs (multi-cubic)}
\State $r_y\!\gets\!\mathsf{rps}_y.r$;\;
       $\alpha\!\gets\!\mathsf{FS}.\mathsf{challenge}(\texttt{"layernorm\_alpha"})$
\State $(\pi^{\gamma\sigma},r_f)\!\gets\!\Prove\text{-}\mathrm{Sumcheck}_3^{\mathrm{multi}}\bigl([\bgamma\!\odot\!\widetilde\bx_{\mathrm{ry}},\widetilde\sigma_{\mathrm{ry}}\!\odot\!\widetilde\by_{\mathrm{ry}}],\alpha,\mathsf{FS}\bigr)$
\State push Hyrax claims into deferred accumulators
$\mathsf{acc}_t,\mathsf{acc}_{td}$ (the model-level batched LN proof groups
LNs by shape and folds all per-LN claims under one extra batching $\eta$)
\State \Return $\{\pi^{\mathrm{mean}},\pi^{\mathrm{sqs}},\pi^{\mathrm{sig}},\pi^{\gamma\sigma},\text{evaluations}\}$
\end{algorithmic}
\end{algorithm}

\subsection{The verifier}

\begin{algorithm}[H]
\caption{$\sys{}.\Verify(\popvk,\piproof,\mathsf{inst},\bx_{\mathrm{in}},\by)$}
\label{alg:verify}
\begin{algorithmic}[1]
\State \textbf{check} $\Com(\bx_{\mathrm{in}})\!=\!\piproof.C_{\bx_{\mathrm{in}}}$
\Comment{public input is bound by re-commitment; the public output is bound below by evaluating $\widetilde\by$ at a transcript-derived random point and chaining through the LM-head proof}
\State $\mathsf{FS}.\mathsf{init}(\texttt{"piformer"})$;\;
       absorb $\piproof.C_{\bx_{\mathrm{in}}}$;\;
       initialise the ${\sim}20$ deferred Hyrax accumulators (one per
       claim-class~/~Hyrax-params group)
\State $C_{\mathrm{cur}}\!\gets\!\piproof.C_{\bx_{\mathrm{in}}}$
\For{$\ell=1,\dots,L$} \Comment{Phase~1 lockstep: absorb all 5 LN IO + aux}
  \State absorb LN$_1$ IO; absorb $C_\bQ,C_\bK,C_\bV$ + (opt.) attn-norm aux
  \State absorb $\LN_Q$ IO ($C_\bQ,C_{\bQ^{\mathrm{n}}}$),
                $\LN_K$ IO ($C_\bK,C_{\bK^{\mathrm{n}}}$), $C_{\bO_{\mathrm{att}}}$
  \State absorb $\LN_{\mathrm{aon}}$ IO; $C_{\bX_{\mathrm{mid}}}\!\gets\!C_{\mathrm{cur}}+p_\ell.C_{\LN_{\mathrm{aon}}.y}$
  \State absorb $\LN_2$ IO; $C_{\mathrm{cur}}\!\gets\!C_{\bX_{\mathrm{mid}}}+p_\ell.C_{\bO_{\mathrm{ffn}}}$
\EndFor
\State $r_{td}\!\gets\!\mathsf{FS}.\mathsf{challenge}(t_{\mathrm{bits}}+d_{\mathrm{bits}})$
\State \textbf{verify} four cross-block sumchecks
$\pi_{qkv},\pi_{op},\pi_{ao},\pi_{ac}$
\State (when normalised) \textbf{verify}
$\pi_{\mathrm{atn}}$ at $r_{\mathrm{norm}}$ and $\pi_Z$
\State \textbf{verify} global FFN Lasso, then $\pi_{ffy},\pi_{ffm}$
\State \textbf{verify} bucketed range batch (one $m_{\mathrm{com}}$ per width)
\State \textbf{verify} model-level batched LN proof (all $5L+1$ LNs)
\State sample $r_{\mathrm{lm}}\!\gets\!\mathsf{FS}.\mathsf{challenge}(t_{\mathrm{bits}}\!+\!v_{\mathrm{bits}})$;\;
       compute $v_{\mathrm{lm}}\!\gets\!\widetilde\by(r_{\mathrm{lm}})$ from the public output
\State \textbf{verify} LM-head sumcheck under the claim
$(r_{\mathrm{lm}},v_{\mathrm{lm}})$;\;
       chain the resulting input-side claim into $\mathsf{acc}_{td}$ against $\piproof.C_{\mathrm{final\_ln\_out}}$
\State drain 13 deferred-accumulator $\mu$-challenges plus 2 read-only $\rho$
fuse challenges; finalise via \texttt{finalize\_many\_with\_mus} (one MSM
pair per Hyrax-params group)
\State \textbf{verify} global intermediate batch open at $r_{td}$
\State \textbf{verify} cross-block matrix-type batch opens
\State (when normalised) \textbf{verify} batch opens at $r_{\mathrm{norm}}$
\State $\textsc{Finalise}(\text{the ${\sim}20$ deferred accumulators})$
\Comment{one MSM pair per group}
\State \textbf{verify} global $\bphi(\bQ^{\mathrm{n}}),\bphi(\bK^{\mathrm{n}})$ Lasso
\State \Return $\mathsf{ACCEPT}$
\end{algorithmic}
\end{algorithm}

%-------------------------------------------------------------------------------
\section{Soundness}
\label{sec:soundness}
%-------------------------------------------------------------------------------

We sketch the soundness analysis; the full proofs follow standard
Schwartz--Zippel and PCS-binding arguments.

\begin{lemma}[Sumcheck soundness]
\label{lem:sumcheck}
For an MLE $f$ of individual degree $\le\delta$, a cheating prover
can make the sumcheck for $H=\sum_{x\in\{0,1\}^n}f(x)$ accept a false
claim $H'\!\neq\!H$ with probability at most $n\delta/|\FF|$.
\end{lemma}

\begin{lemma}[Hyrax binding]
\label{lem:hyrax}
Under the discrete-log assumption on $\GG$, the Hyrax PCS is
computationally binding: no PPT adversary can open a single commitment
to two different evaluations at the same point with non-negligible
advantage.
\end{lemma}

\begin{theorem}[Cross-block batching is sound]
\label{thm:cross-block}
Let $\textsc{CrossBatch}$ (Algorithm~\ref{alg:cross-block}) reduce
$L$ per-block claims $\{H_\ell\}$ to one batched claim
$H^\star\!=\!\sum_\ell \eta^{\ell-1}H_\ell$, and let the inner sumcheck
be sound with error $\le n\delta/|\FF|$. Then $\textsc{CrossBatch}$
is sound with total error
\[
  \varepsilon_{\mathrm{batch}}\;\leq\;\frac{L-1}{|\FF|}+\frac{n\delta}{|\FF|}.
\]
\end{theorem}
\begin{proof}
The challenge $\eta$ is drawn from $\mathsf{FS}$ \emph{after} every
per-block claim and per-block witness commitment is absorbed. The
prover is therefore committed to $\{H_\ell\}$ before learning $\eta$.
If any single claim is false, $\hat H_{\ell^\star}\!\neq\!H_{\ell^\star}$,
then $P(\eta)=\sum_\ell \eta^{\ell-1}(\hat H_\ell-H_\ell)$ is a
non-zero polynomial in $\eta$ of degree $\le L-1$. By Schwartz--Zippel,
$P(\eta)=0$ holds for at most $L-1$ values of $\eta$, so a uniformly
random $\eta$ catches the false claim with probability
$\geq 1-(L-1)/|\FF|$. Conditional on the combined claim being honest,
the inner sumcheck has soundness $n\delta/|\FF|$. A union bound gives
the stated error.
\end{proof}

\begin{theorem}[Homomorphic residuals]
\label{thm:residual}
Let $C_\bA\!=\!\Com(\bA)$, $C_\bB\!=\!\Com(\bB)$. The verifier-computed
commitment $C_\bA+C_\bB$ binds $\bA+\bB$ unconditionally given Hyrax
binding (Lemma~\ref{lem:hyrax}).
\end{theorem}
\begin{proof}
Hyrax commits row-by-row as MSMs over public generators
$\mathbf g$. By bilinearity of MSM,
$C_\bA+C_\bB=(\mathrm{MSM}(\mathbf g,A_i)+\mathrm{MSM}(\mathbf g,B_i))_i
=(\mathrm{MSM}(\mathbf g,A_i+B_i))_i=\Com(\bA+\bB)$. Binding of
$\bA+\bB$ then follows from binding of $\bA$ and $\bB$.
\end{proof}

\begin{theorem}[Globally batched range proof is sound]
\label{thm:range}
Sharing one $m_{\mathrm{com}}$ per width bucket across all range witnesses
in the bucket (LN $\sigma$/$y$ across all blocks plus the final LN, plus
the optional attention-normalisation residuals) is sound: a cheating
prover cannot supply chunk values outside $[0,2^{16})$ for any single
witness. Moreover, the zerocheck-folded inverse polynomial binds $h$ to
$1/(\alpha-V)$ on the boolean hypercube without committing $h$, with
soundness error $O(n/|\FF|)$ per bucket where $n$ is the bucket variable
count.
\end{theorem}
\begin{proof}
Each bucket is processed independently and the argument applies per bucket.

\emph{Chunk-value binding.} The transcript-ordering invariant is that
$m_{\mathrm{com}}$ is absorbed before any per-witness sumcheck challenge
$r_v^{(b)}$ in the bucket is sampled. Since $m$ aggregates all chunk
occurrences across all witnesses and all
$C\!=\!\lceil\mathrm{bits}/16\rceil$ chunks per witness, the prover must
commit to $m$ before seeing any sumcheck challenge. If the prover
supplied a chunk value $v^*\!\geq\!2^{16}$, the LogUp
identity~\cite{habo2024logup} would require
$m[v^*]\!\cdot\!T_{\mathrm{id}}[v^*]$ to balance, but $T_{\mathrm{id}}$
has no entry for $v^*$, contradicting the committed $m$
(Lemma~\ref{lem:hyrax}).

\emph{Inverse-polynomial binding (zerocheck fold).} Let
$\tilde h$ be a (possibly malicious) prover's choice of inverse-polynomial
MLE, $V$ the (committed) chunk MLE, and define
$\tilde s(x)\!:=\!\tilde h(x)(\alpha-V(x))\!-\!1$. The bucket sumcheck
proves
\[
  \sum_x\!\bigl[\,w(x)\tilde h(x)(1+\beta(\alpha\!-\!V(x)))
   + \gamma_{\mathrm{fold}}\!\cdot\!\mathrm{eq}(r_{zc},x)\tilde s(x)\bigr]=\Lambda,
\]
where $\Lambda$ is computed from the prover-supplied per-chunk sums and
absorbed into the transcript \emph{before} $\gamma_{\mathrm{fold}}$ and
$r_{zc}$ are sampled. If $\tilde h\!\neq\!1/(\alpha\!-\!V)$ on the
boolean hypercube, then $\tilde s$ is non-zero on at least one boolean
point, hence non-zero as a polynomial of total degree $\le\!2n$, and by
Schwartz--Zippel
$\Pr_{r_{zc}}[\tilde s(r_{zc})\!=\!0]\!\le\!2n/|\FF|$. Conditioning on
$\tilde s(r_{zc})\!\neq\!0$, the true integrand value is linear in
$\gamma_{\mathrm{fold}}$ with non-zero slope, so the probability it
equals $\Lambda$ over a fresh $\gamma_{\mathrm{fold}}$ is at most
$1/|\FF|$. If the sum differs from $\Lambda$, standard sumcheck
soundness rejects with probability $\ge 1\!-\!3n/|\FF|$. Union-bounding
yields $\le 6n/|\FF|$ soundness error per bucket. The pair-padding
positions are filled with $\tilde h\!=\!\alpha^{-1}$, on which both the
main term ($w\!=\!0$) and the zerocheck term ($\alpha^{-1}\alpha\!-\!1\!=\!0$)
vanish, so padding contributes nothing to either side of the identity.

Once $\tilde h$ is forced to the unique inverse on the hypercube, the
per-chunk claims $\sum_i\tilde h_p[i]$ recover the true LogUp LHS, and
the model-level grand-sum check
$\sum_p\mathrm{claim}_p\!=\!\mathrm{rhs\_claim}$ closes the argument
against the multiplicity-table side.
\end{proof}

\begin{theorem}[Model-level batched LayerNorm is sound]
\label{thm:lnbatch}
The single $\textsc{ProveLNsBatched}$ call covering all $5L+1$ LNs in the
model (per block: $\LN_1,\LN_Q,\LN_K,\LN_{\mathrm{aon}},\LN_2$; plus the
final LN) is as sound as $5L+1$ independent LN sub-proofs run with
independent challenges.
\end{theorem}
\begin{proof}
The batched LN proof folds the $5L+1$ per-LN claims via Fiat--Shamir
powers of an independent batching challenge $\eta$, drawn after every
per-LN IO commitment ($\bx,\by,\mathrm{sum\_x},\mathrm{sq\_sum},\sigma$)
has been absorbed into the transcript and after the bucketed range batch
has committed to all $\sigma$/$y$ chunks. The same Schwartz--Zippel
argument as Theorem~\ref{thm:cross-block} applies (with $L$ replaced by
$5L+1$): a single false LN claim makes the combined polynomial in $\eta$
non-zero of degree $\le 5L$, caught with probability $\ge 1-5L/|\FF|$.
The shared inner sumchecks (mean, variance, Y-fusion) inherit
Lemma~\ref{lem:sumcheck} soundness on the combined claim.
\end{proof}

\begin{theorem}[Attention-normalisation sumcheck is sound]
\label{thm:attnnorm}
When normalised attention is enabled, the cubic multi-batched sumcheck
for $S\!\cdot\!\bN-\mathrm{rem}-Z\!\cdot\!\bY^{\mathrm{att}}=0$ and
$\lambda(Z-1-\mathrm{rem}-\mathrm{diff})=0$, combined with the range proofs
on $\mathrm{rem},\mathrm{diff}$ (Theorem~\ref{thm:range}) and the auxiliary
$Z$ sumcheck, soundly enforces the floor-division identity
$\bY^{\mathrm{att}}_{tj}=\lfloor S\!\cdot\!\bN_{tj}/Z_t\rfloor$ at every
$(t,j)$.
\end{theorem}
\begin{proof}[Proof sketch]
Range proofs constrain $\mathrm{rem}\!\geq\!0$ and $\mathrm{diff}\!\geq\!0$.
The second identity then gives $\mathrm{rem}\!\leq\!Z\!-\!1$, i.e.\
$0\!\leq\!\mathrm{rem}\!<\!Z$. The first identity expresses
$Z\!\cdot\!\bY^{\mathrm{att}}=S\!\cdot\!\bN-\mathrm{rem}$ with the unique
quotient given $0\!\leq\!\mathrm{rem}\!<\!Z$. Both identities are checked
at one Fiat--Shamir random point of $t_{\mathrm{bits}}\!+\!d_{\mathrm{bits}}$
variables with cubic-sumcheck error and a $\lambda$-batching error of
$O(1)/|\FF|$. The auxiliary $Z$ sumcheck binds $Z_t$ to
$\bphi(\bq_t^{\mathrm{n}})\!\cdot\!\sum_s\bphi(\bk_s^{\mathrm{n}})$ via the
already-committed $\bphi(\bQ^{\mathrm{n}}),\bphi(\bK^{\mathrm{n}})$ MLEs.
Hyrax binding closes the loop.
\end{proof}

\begin{theorem}[\sys{} end-to-end soundness]
\label{thm:end-to-end}
For any PPT adversary $\calA$, the probability that
$\Verify(\popvk,\piproof,\mathsf{inst},\bx_{\mathrm{in}},\by)=\mathsf{ACCEPT}$
but $\calM_\theta(\bx_{\mathrm{in}})\!\neq\!\by$ is at most
\[
  \varepsilon_{\mathrm{sound}}\;\leq\;
  \frac{m_{\mathrm{sc}}\,n_{\max}\delta_{\max}+m_{\mathrm{batch}}(L-1)+m_{\mathrm{open}}}{|\FF|}
  +\varepsilon_{\mathrm{DL}},
\]
where $m_{\mathrm{sc}}$ is the number of sumcheck invocations after
batching, $m_{\mathrm{batch}}\!=\!6$ is the number of cross-block
batches, $m_{\mathrm{open}}$ covers the random-linear-combination
checks used in batched openings and lookups, $n_{\max}$ is the
maximum number of sumcheck variables, $\delta_{\max}\!\leq\!3$ is the
maximum round-polynomial degree, and $\varepsilon_{\mathrm{DL}}$ is
the DL advantage of $\calA$ on $\GG$.
\end{theorem}
\begin{proof}[Proof sketch]
By hybrid argument over the $L$ blocks. (i) By Lemma~\ref{lem:hyrax},
any forged Hyrax opening yields a DL solution. (ii) Block-level chaining is
sound by Theorem~\ref{thm:residual}, with the sandwich-norm residual
$C_{\bX_{\mathrm{mid}}}=C_{\bx_{\mathrm{in}}}+C_{\LN_{\mathrm{aon}}.y}$
inheriting binding from $C_{\LN_{\mathrm{aon}}.y}$. (iii) Within each block,
the bucketed global range batch over all witnesses is sound by
Theorem~\ref{thm:range}; the model-level batched LN proof covering all
$5L\!+\!1$ LNs is sound by Theorem~\ref{thm:lnbatch}; when normalised
attention is enabled, the floor-division identity is sound by
Theorem~\ref{thm:attnnorm}; the cross-block sumchecks are sound by
Theorem~\ref{thm:cross-block}; the committed-output Lasso is sound by
Lemma~\ref{lem:hyrax} plus Lemma~\ref{lem:sumcheck}. The Lasso
index-binding step ties the lookup queries to the post-QK-norm tensors
$C_{\bQ^{\mathrm{n}}},C_{\bK^{\mathrm{n}}}$, which are themselves bound to
$C_\bQ,C_\bK$ through the LN-batched proof for $\LN_Q,\LN_K$. (iv) The
final layer plus LM head reduce to one LN plus one projection. A union
bound over all sub-protocols gives the bound. For typical parameters
($L\!\le\!12$, $n_{\max}\!\le\!64$, $\delta_{\max}\!\le\!3$) over BN254,
the statistical term is below $2^{-240}$ and is dominated by the DL term.
\end{proof}

\subsection{Zero knowledge}
The current implementation is \emph{not} zero-knowledge: weight
commitments are public VK entries, and intermediate evaluation claims
(e.g., $\bX_{\mathrm{n1}}(r_t,r_d)$) are revealed in the sumcheck
transcripts. Standard techniques apply: blinding all witness
polynomials with random masking terms before commitment, and using
zero-knowledge variants of sumcheck and Hyrax. We discuss this
further in \S\ref{sec:disc}.

%-------------------------------------------------------------------------------
\section{Implementation}
\label{sec:impl}
%-------------------------------------------------------------------------------

We implement \sys{} in $\sim$19\,k lines of Rust over BN254 plus a
PyTorch training pipeline. The Rust prover/verifier and CLI are
organised as in Table~\ref{tab:loc}.

\begin{table}[t]
\centering
\caption{\sys{} Rust implementation, lines of code per module.}
\label{tab:loc}
\small
\begin{tabular}{lr}
  \toprule
  \textbf{Module} & \textbf{LoC} \\
  \midrule
  \texttt{poly/} (dense MLE, equality, helpers)             & $\sim$0.6\,k \\
  \texttt{pcs/} (Hyrax PCS, batch accumulator)              & $\sim$1.3\,k \\
  \texttt{subprotocols/} (sumcheck $\times3$, GKR combine)  & $\sim$1.5\,k \\
  \texttt{lookup/} (Lasso multi, range batched, quant.)     & $\sim$3.4\,k \\
  \texttt{attention/} (LN, projection, LLA, ternary check)  & $\sim$4.6\,k \\
  \texttt{ffn/} (FFN sub-prover)                            & $\sim$0.7\,k \\
  \texttt{cross\_layer/} (cross-layer projection sumcheck)  & $\sim$0.6\,k \\
  \texttt{prover.rs}, \texttt{verifier.rs}, \texttt{setup.rs} & $\sim$6.0\,k \\
  \texttt{bin/piformer/} (CLI, codecs, sample)              & $\sim$0.5\,k \\
  \midrule
  \textbf{Total}                                             & $\sim$19\,k \\
  \bottomrule
\end{tabular}
\end{table}

\paragraph{PyTorch pipeline.}
The training pipeline implements \texttt{TernaryLinear} (with STE),
\texttt{StructuredLookupActivation} ($\bphi$ with sub-table forward
plus STE backward), and \texttt{LinearAttentionLayer}. The integer
forward pass used to construct the witness is bit-identical to the
Rust circuit, so a witness exported from PyTorch is accepted by the
Rust prover with no quantisation gap. The exporter also emits the
Lasso instances ($\bphi(\bQ),\bphi(\bK),\bphi(\bM)$ tables, query
indices, and outputs) directly from the same integer arithmetic.

\paragraph{Single-head, causal supported, no ZK.}
The current Rust prover supports single-head attention with optional
end-to-end causal masking (selected at setup; the prover swaps in three
cubic-multi-batched sub-protocols whose third multiplicand absorbs the
causal indicator polynomial) and no zero-knowledge masking layer. Multi-head
splits the attention sumchecks into parallel batches with the same protocol
structure, and ZK is an additional blinding layer over committed witnesses
(\S\ref{sec:disc}).

%-------------------------------------------------------------------------------
\section{Discussion and Limitations}
\label{sec:disc}
%-------------------------------------------------------------------------------

\paragraph{What is and is not in the SNARK.}
The Rust prover proves the full normalised LLA output
$\bY^{\mathrm{att}}=\lfloor S\!\cdot\!\bphi(\bQ^{\mathrm{n}})(\bphi(\bK^{\mathrm{n}})^\top\bV)/Z\rfloor$
followed by the ternary output projection
$\bO_{\mathrm{att}}=\bY^{\mathrm{att}}\bW_O+\bb_O$ and the sandwich-norm
$\LN_{\mathrm{aon}}(\bO_{\mathrm{att}})$ before the residual sum. The
Python pipeline supports two attention modes:
\texttt{normalized\_fixed} (proven end-to-end by the Rust circuit,
including the floor division) and \texttt{prover} (un-normalised numerator
only, retained for legacy export); \texttt{normalized\_float} is rejected
at export time. \emph{Causal masking is fully supported} end-to-end: when
\texttt{causal=true} the prover swaps in cubic-multi-batched
\texttt{batch\_attn\_out\_causal} / \texttt{batch\_attn\_ctx\_causal} /
\texttt{attn\_z\_causal\_sumcheck} sub-protocols that fold the causal
indicator polynomial into the third multiplicand, so no separate
prefix-context commitment is needed. The remaining limitations are
multi-head attention (the prover currently rejects $n_{\mathrm{heads}}\neq 1$
at setup) and zero-knowledge masking. Neither changes the protocol
architecture: multi-head splits the attention sumchecks into parallel
batches, and ZK is an additional blinding layer over committed witnesses.

\paragraph{Model-quality evaluation.}
Because \sys{} changes the attention kernel and constrains weights to
ternary values, a deployment-quality comparison must train matched
baselines under the same parameter budget, context length, dataset,
and quantisation regime, then report both task quality and proof
cost. We leave this to future work and present \sys{} here as
evidence that the co-designed architecture is efficiently provable.

\paragraph{From prototype to deployment.}
Our reference implementation is CPU-only and single-threaded. GPU
MSM and parallel sumcheck~\cite{qu2025zkgpt} would translate
directly to \sys{}: every dominant inner loop (cross-block batched
sumcheck, Lasso multi-instance, batched Hyrax open) is data-parallel,
and the global Hyrax accumulators are pure MSMs at the boundary.
Streaming token generation is the natural fit for Nova-style
folding~\cite{kothapalli2022nova}: the running context matrix
$\bM=\bphi(\bK)^\top\bV$ is the IVC accumulator, and one folding step
per token yields constant amortised proving cost.

\paragraph{Beyond \sys{}.}
The Rust codebase already contains two stand-alone sub-protocols that
are not yet wired into the end-to-end prover. The first is an
in-circuit ternary-weight check
(\texttt{attention/ternary\_check.rs}) via a 4-element Lasso table
$[0,1,r-1,0]$; folding this into setup yields a verifying key that
carries a binding proof $\bW\!\in\!\calT^{m\times n}$. The second is
a single cubic sumcheck (\texttt{cross\_layer/projection.rs}) that
collapses $L$ per-layer projection sumchecks into one of length
$\log L+\log d_{\mathrm{in}}$ by introducing a layer-index variable;
this gives a further factor of $L$ saving on the projection
transcript. Together with an EVM-targeted Hyrax verifier and a
zero-knowledge masking layer, these lay out the path from a research
prototype to an on-chain verifiable inference service.

%-------------------------------------------------------------------------------
\section{Related Work}
\label{sec:related}
%-------------------------------------------------------------------------------

\paragraph{Verifiable ML.}
zkCNN~\cite{liu2021zkcnn} gave the first GKR-based proof system for
neural network inference, focused on convolutional networks.
\cite{abbaszadeh2024zero} extends ZK to proofs of \emph{training}
for deep networks. zkLLM~\cite{sun2024zkllm} adapts a custom IOP to
softmax LLM inference using \texttt{tlookup}-style structured lookups,
with \texttt{zkAttn} as a tailored attention sub-protocol;
zkGPT~\cite{qu2025zkgpt} uses a Plonky2 backend with constraint
fusion across operations of a transformer block. Both target the
standard softmax + GeLU + LayerNorm stack and pay a significant cost
for every non-arithmetic operation. \sys{} departs by co-designing
the model and proof system: every operation is either a sumcheck or
a Lasso lookup by construction, and the model is allowed to recover
expressivity through learnable lookup tables that are free under
Lasso.

\paragraph{Sumcheck and lookup arguments.}
The sumcheck protocol underlies modern transparent
SNARKs~\cite{setty2020spartan,thaler2022proofs}. Lasso provides a
practical lookup argument with sub-linear prover cost on
structured-decomposable tables~\cite{setty2024unlocking,arun2024jolt,fomenko2025understanding},
and is the natural fit for the additive sub-table structure of
\sys{}'s $\bphi$. LogUp-style multiplicity arguments~\cite{habo2024logup}
back the globally batched range proof.
Hyrax~\cite{wahby2018doubly} gives a transparent, discrete-log-based
polynomial commitment with $O(\sqrt N)$ proofs.

\paragraph{Linear attention and ternary networks.}
Linear attention~\cite{katharopoulos2020transformers} replaces
softmax with a feature-map kernel and recovers the
$O(Td^2)$-vs-$O(T^2 d)$ complexity trade-off;
\cite{wang2020linformer} projects the keys and values to a low-rank
space; Reformer~\cite{kitaev2020reformer} uses LSH for sparsity.
BitNet b1.58~\cite{ma2025bitnet} introduced ternary weights as a
replacement for FP16 dense layers in LLMs, motivated by inference
energy on edge devices; \sys{} re-purposes the same primitive for
SNARK proving cost.

%-------------------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}
%-------------------------------------------------------------------------------

We presented \sys{}, a SNARK for verifiable transformer inference in
which the model is made SNARK-friendly so the proof system has nothing
expensive to absorb. Three architectural changes drive the design:
softmax becomes a kernel attention with a learnable additive-lookup
feature map, every projection becomes ternary so its sumcheck
contraction has no field multiplications, and LayerNorm becomes an
integer reformulation provable through sumcheck and chunked range
proofs. The proof system batches per-block sub-protocols into six
model-level sumchecks, two global Lasso proofs, and one global range
proof per block, with all final Hyrax openings flowing through ten
deferred batch accumulators. We implemented the system in $\sim$19\,k
lines of Rust over BN254 and gave formal cross-block batching
soundness theorems. We leave a deployment-quality model evaluation
and the engineering steps (multi-head, ZK masking, GPU MSM, on-chain
verifier) to future work; we believe \sys{} provides a credible
foundation for production-grade verifiable transformer inference.

%-------------------------------------------------------------------------------
\bibliographystyle{plain}
\bibliography{ref}
%-------------------------------------------------------------------------------

\end{document}
